Mettl5 coordinates protein production and degradation of PERIOD to regulate sleep in Drosophila

Version 1

Abstract

Summary

Sleep is crucial for animal physiology, primarily governed by the brain, and disruptions are prevalent in various brain disorders. Mettl5, associated with intellectual disability (ID) often accompanied by sleep disturbances, remains poorly understood in its role causing these disorders. Previous research demonstrated that Mettl5 forms a complex with Trmt112, influencing rRNA methylation. In our study, we explored sleep phenotypes due to Drosophila Mettl5 mutations. Rescue experiments pinpointed Mettl5’s predominant role in neurons and glia marked by Mettl5-Gal4 in sleep regulation. Notably, a Trmt112 mutation mirrored these sleep disturbances, implicating translational regulation via the Mettl5/Trmt112 complex. Subsequent RNA-seq and Ribo-seq analyses unveiled downstream events from Mettl51bp mutations, revealing altered expression levels of proteasome components and Clock genes. Rescue experiments confirmed that the net increased PERIOD protein is responsible for the sleep phenotype. This investigation sheds light on ribosome, clock genes, and proteasome interplay in sleep regulation, underscoring protein synthesis and degradation’s integrative role. These findings could potentially provide an example of in vivo study of the function of rRNA methylation, expand our understanding of the role of protein homeostasis in sleep and inspire explanations on the ID related sleep phenotypes.

Introduction

Sleep serves critical functions for animals. Investigating the molecular mechanisms underlying sleep holds profound importance for both fundamental research and practical applications. While previous studies have identified several sleep regulators (Du et al., 2021), gaps persist in comprehending fundamental cellular processes like protein synthesis and degradation in relation to sleep.

Evidence suggests a correlation between sleep regulation and protein synthesis and degradation, both of which control protein levels dynamically. Protein synthesis is particularly active during sleep (Lyons et al., 2023). Sleep deprivation targets translational initiation, impacting this process (Costa et al., 2019). The proteasome, crucial for ubiquitination-mediated protein degradation, plays a key role in synapse homeostasis by modulating ribosomal components (Costa et al., 2019), thereby influencing synaptic function. Additionally, proteasome components exhibit a circadian expression pattern at the transcriptional level in human cells (Desvergne et al., 2016). A recent study in Drosophila showed that mutation of a proteasome component alters sleep patterns (Fernández et al., 2020). The roles of protein synthesis and degradation in sleep regulation, as well as their interactions, remain open questions.

Several factors influencing proteasome assembly and subunit transcription have been identified (Kapetanou et al., 2022). However, there remains a significant gap in research regarding the translational control of proteasome subunits. Furthermore, it has been reported that proteasome inhibition can lead to changes in ribosome function (Galimberti et al., 2016; Costa et sl., 2019; Palanca et al., 2014). An open question persists as to whether the proteasome is regulated by the status of the translation machinery. Exploring this relationship could provide deeper insights into the interplay between proteasome function and translation.

Drosophila Mettl5 exhibits 18S ribosomal RNA m6A methyltransferase activity (Leismann et al., 2020). In both Drosophila and human, Mettl5 interacts with Trmt112 to facilitate this function (Leismann et al., 2020; van Tran et al., 2019) Depletion of Mettl5 results in the absence of m6A modification on 18S rRNA, yet does not affect rRNA maturation. The loss of Mettl5-mediated m6A modification influences fly orientation. In mammalian cells, this gene was found mediating 18S rRNA N6-methyladenosine (m6A) modification and controlling stem cell fate determination and neural function (Wang et al., 2020). However, it remains intriguing whether Mettl5 impacts global or transcript-specific translation profile, especially in tissues in vivo instead of cell lines. Furthermore, Mettl5, associated with intellectual disability (ID) often accompanied by sleep disturbances, remains poorly understood in its role causing these disorders. Thus, test whether the loss of Mettl5 loss affect behavior such as sleep in Drosophila would shed light on this problem.

We discovered that Mettl5 regulates sleep. To investigate the underlying mechanism, we used RNA-seq and Ribo-seq to reveal downstream events associated with Mettl5 mutations. We identified multiple clock genes and proteasome components regulated by Mettl5, suggesting a coordinated regulation of protein production and degradation, crucial for protein homeostasis. Further experiments showed that the protein level of Period was upregulated, contributing to the sleep phenotype. This study highlights how ribosome defects can impact proteasome function, demonstrating a mechanism that coordinates protein degradation with protein production. Additionally, by mapping the genome-wide profile of an rRNA methylation modifier, this study offers insights into the gene-specific roles of ribosome function.

Results

Mettl5 is a regulator of Drosophila sleep

Mettl5 is predicted to have a N6 adenine specific nucleic acids methylase domain. In order to study the function of this gene in sleep regulation, we generated mutants by CRISPR-Cas9 deletion. Allele Mettl51bp, which deleted 1bp in the coding region results in a truncated version (Figure 1A) and down regulation of Mettl5 at mRNA level (Figure 1B). Another allele Mettl59bp, which results in 3 amino acids deleted version of Mettl5 (Figure 1A) does not cause significant change at mRNA level (Figure 1B). We found that the heterozygous Mettl51bp mutation results in decreased total sleep during the night time, which is resulted from the effect of decrease of sleep at early night time (Figure 1C-F). The sleep curve showed a significant down regulation of sleep at ZT12 to ZT16 (Figure 1C). The calculation of the awake percentage at ZT14 indicated a significant increase of this parameter in Mettl51bp (Figure 1G). In addition, Mettl5 expression was significantly up regulated during the sleep recovery period after mechanical sleep deprivation (Figure 1M-N). Consistently, the sleep recovery was down regulated in the Mettl5 mutant (Figure 1O), indicating the sleep homeostasis was affected by Mettl5. In conclusion, the truncated mutation of Mettl5 resulted in sleep phenotypes.

Mettl5 is a regulator of Drosophila sleep.

(A) Diagram illustrating CRISPR-Cas knockout of 1 or 9 bases in the Mettl5 gene. (B) Relative expression of Mettl5 mRNA in Mettl51bp and Mettl59bp mutant male flies compared to control flies. (C) Sleep curve throughout the day for Mettl5 mutant male flies and control flies. (D) Total sleep of Mettl5 mutant male flies and control flies. (E) Sleep bout duration of Mettl5 mutant male flies and control flies. (F) Number of sleep bouts of Mettl5 mutant male flies and control flies. (G) Percentage of awake for Mettl5 mutant flies and control flies. (H) Sleep curve throughout the day for Mettl5 mutant male flies, induced Mettl5 over expression male flies and control flies. (I) Total sleep of Mettl5 mutant male flies, induced Mettl5 over expression male flies and control flies. (J) Sleep bout duration of Mettl5 mutant male flies, induced Mettl5 over expression male flies and control flies. (K) Number of sleep bouts of Mettl5 mutant male flies, induced Mettl5 over expression male flies and control flies. (L) Percentage of awake for Mettl5 mutant flies, Mettl5 over expression male flies and control flies. (M) Sleep curve is tracked throughout the entire day prior to sleep deprivation and during the daytime sleep rebound period. (N) Mettl5 mRNA expression level at different time points. W(wake), SD (sleep deprivation), SR (sleep recovery). (O) The daytime sleep recovery during the six hours following sleep deprivation in Mettl5 mutant male flies. For * stands for p<0.05, ** stands for p<0.01, *** stands for p<0.001, ns stands for not significant.

To verify the specificity of the sleep phenotype caused by Mettl51bp, we did the rescue experiment. The results showed that supplying of one copy of Mettl5 to Mettl51bprescued the sleep phenotype in sleep amount and awake percentage at ZT14 (Figure 1H-L). These results verified the specificity of the phenotype.

To verify in which cells did Mettl5 function for the sleep phenotype, we checked the expression pattern of Mettl5-Gal4. The results indicated that a portion of the Mettl5-Gal4 marked cells colocalized with Repo antibody staining (Figure S1A-C). Another portion of the Mettl5-Gal4 marked cells colocalized with ELAV antibody staining (Figure S1D-F). In addition, we also tested the climbing ability of Mettl5 mutation which showed an increase (Figure S1G). The sleep arousal assay also indicated an increased sleep arousal (Figure S1H). Moreover, the lifespan increased in the mutant (Figure S1I).

Mettl5 regulation on Drosophila sleep was dependent on its methyltransferase activity

Previous studies have shown that Mettl5 interacts with Trmt112 to regulate 18S rRNA m6A modification (Leismann et al., 2020; van Tran et al., 2019). Consistently, we found that the methylation level in both total and 18S rRNA decreased in heterozygous Mettl51bp. We tested the m6A level using LC-MS/MS experiment. The m6A level decreased in heterozygous Mettl51bp either in both total RNA and 18S rRNA (Figure 2A-B). To investigate the mechanism of Mettl5 regulation on sleep. We examined the phenotype of Trmt112 knockdown. We found that the Mettl5-Gal4 driven Trmt112 RNAi resulted in a similar phenotype to Mettl51bp (Figure 2C-G), suggesting that Mettl5 regulates sleep through its 18S rRNA m6A modification activity. The efficiency of the Trmt112 RNAi line was tested in previous study (López-Varea et al., 2021). More important, we performed a rescue experiment with a mutated form of Mettl5 that lacks the NPPF amino acids required for its methyltransferase activity. We observed that this mutant Mettl5 failed to rescue the sleep phenotype caused by Mettl51bp (Figure 2H-L). These results collectively indicated that the methyltransferase activity of Mettl5 is essential for its role in Drosophila sleep.

Mettl5 regulation of Drosophila sleep was dependent on its methyltransferase activity.

(A) The m6A level in the total RNA of Mettl5 mutant male flies. (B) The m6A level in the 18S rRNA of Mettl5 mutant male flies. (C) The sleep curve throughout the day shows the sleep pattern of induced Trmt112 RNAi male flies and control flies. (D) Total sleep of induced Trmt112 RNAi male flies and control flies. (E) Sleep bout duration in induced Trmt112 RNAi male flies and control flies. (F) Number of sleep bouts in induced Trmt112 RNAi male flies and control flies. (G) Percentage of awake in Trmt112 RNAi and control flies. (H) Sleep curve throughout the day for Mettl5 mutant male flies, induced Mettl5m over expression male flies and control flies. (I) Total sleep of Mettl5 mutant male flies, induced Mettl5m over expression male flies and control flies. (J) Sleep bout duration in Mettl5 mutant male flies, induced Mettl5 over expression male flies and control flies. (K) Number of sleep bouts in Mettl5 mutant male flies, induced Mettl5 over expression male flies and control flies. (L) Percentage of awake in Mettl5 mutant male flies, induced Mettl5 over expression male flies and control flies. For * stands for p<0.05, ** stands for p<0.01, *** stands for p<0.001, ns stands for not significant.

RNA-seq and Ribo-seq revealed the downstream gene profile of Mettl5

To better understand the downstream events of Mettl5, we analyzed RNA-seq and Ribo-seq data to evaluate the effects of Mettl51bpon gene expression. The analysis of the RNA-seq and Ribo-seq data followed by the Principal Coordinates Analysis (PCoA) showed a clear separation between the two groups (Figure S2A, B, D, E). High correlation was found between the three biological repeats for each sample (Figure S2C, F), indicating that RNA-seq and Ribo-seq quality was high. Differentially expressed gene analysis of RNA-seq showed that 1053 genes (|log2 (Fold Change) | ≥ 1 & padj< 0.05) were significantly changed at the transcription level (Figure 3A), including 217 up-regulated and 836 down-regulated genes (Figure 3B, Table S1). Differentially expressed gene analysis of Ribo-seq showed that 258 genes were significantly changed at the translation level (|log2 (Fold Change) | ≥ 0.265 & padj< 0.05), including 149 up-regulated and 150 down-regulated genes (Figure 3C, Table S1). The top 100 differentially expressed genes found in RNA-seq (Figure S3A) and Ribo-seq (Figure S3B) were shown in a heatmap generated according to the rlog transformed values.

RNA-seq and Ribo-seq analysis revealed changes in the gene profile of Mettl51bp.

(A) Venn diagram depicting the number of significant differentially expressed genes revealed by RNA-seq and Ribo-seq. (B) Volcano plot representing the differentially expressed genes identified by RNA-seq. Genes that met the criteria of |log2(fold-change)| ≥ 1 and p.adjust < 0.05 were considered significantly expressed, marked in orange for downregulation and green for upregulation. (C) Volcano plot representing the differentially expressed genes identified by Ribo-seq. Candidates that satisfied the criteria of |log 2 (fold-change) | ≥ 0.265 and p.adjust <0.05 were regarded as significantly expressed, marked in red for downregulation and blue for upregulation, respectively. (D-K) Gene set enrichment analysis of differentially expressed genes revealed by RNA-seq (D-G) and Ribo-seq (H-K). All the plots are generated using the KEGG gene set database. The bar chart at the bottom of each panel shows the distribution of target genes for each pathway according to their rank position. Each vertical line represents a gene. Genes on the left show positive correlation with Mettl51bp, while, genes on the right show negative correlation with Mettl51bp. The green line indicates the enrichment score (ES), and NES stands for normalized enrichment score. (L) Distribution of the differentially expressed genes revealed by both RNA-seq and Ribo-seq. (M) Cumulative distribution of TE frequencies among w1118 and Mettl51bp (N, O) GO and KEGG enrichment of significantly changed TE-related genes between w1118 and Mettl51bp. The color of the bar indicates the enrichment p.adjust value. (P) KEGG network showing the top 5 pathway and associated genes. The size of the dots represents the number of genes in the pathway.

To explore the biological implication of the differentially expressed genes at transcriptional and translational levels, Gene ontology (GO) enrichment was performed. At the transcriptional level (Figure S4A, C), the most significantly enriched biological processes GO terms were related to cellular response, biosynthetic process and metabolic process, such as cellular response to chemical stimulus, small molecule biosynthetic process, fatty acid metabolic process (Figure S4A). Several term related gene network also be shown (Figure S4C). At the translation level (Figure S4B, D), the most significantly enriched biological processes GO terms were related to organic acid metabolic process, amino acid metabolic process, lipid catabolic process, cellular respiration, and transport mechanisms. Additionally, we found prominent terms related to circadian rhythm genes, including circadian regulation of gene expression and entrainment of the circadian clock (Figure S4B, D).

The GO results usually contain a long list of enriched terms that have highly redundant information and are difficult to summarize. So, we performed a simplify Enrichment analysis that visualizes the summaries of clusters by word cloud for the GO enrichment result at transcriptional and translational levels. We found that the main biological process was metabolic, the response signaling stimulus; the development of wing disc; imaginal et.al system; the homeostasis and transport for ion-channel; the cycle of cytokinesis and transition in transcriptional significant DEG (Figure S5A). Except for the above main biological progress, the significant translated genes also showed enrichment for the sleep wake, cycle related behavior circadian rhythm and neuron death (Figure S5D). For cellular components, the main enrichments are chromosome structures, complexes, membranes, and vesicles (Figure S5B). Molecular functions primarily include enzymatic activities, ion and protein binding, receptor activities, and transcription processes (Figure S5C). The RNA-seq results are very similar to the Ribo-seq results (Figure S5E-F).

Gene Set Enrichment Analysis (GSEA) revealed pathways that were enriched in differentially expressed genes at both transcriptional and translational levels. As the significant genes enrichment analysis will lose some key important message which are caused by unreasonable screening parameters, these gene sets ranked by log2fold-change only (not screened by p values) were considered in the GSEA. The GSEA-GO and GSEA-KEGG enrichment results of the transcriptional and translational genes were sorted according to the Normalized enrichment score (NES) value (Figure S6A-D). We found some important pathway such as proteasome pathway that were suppressed at both transcriptional and translational levels (P.adjust<0.05) (Figure 3E, I). Arginine and proline metabolism and insect hormone biosynthesis were significantly affected at transcriptional level (P.adjust<0.05) (Figure 3D, G). While, fatty acid degradation, Phosphonate and Phosphinate metabolism,etc. were significantly affected at translational level (P.adjust<0.05) (Figure 3H-K).

In particular, GSEA analysis showed some interesting results. Gene ontology analysis (GSEA-GO) of the GSEA gene set showed a suppression of mitochondria related activities both at transcriptional and translational levels (Figure S6A, C). The aromatic amino acid metabolic and catabolic process were activated at transcriptional level (Figure S6A). While, the signaling receptor, molecular transducer activity related term was activated at translational level (Figure S6C). The KEGG analysis (GSEA-kegg) indicated that at transcriptional level, we found that the pathway Pentose and glucuronate interconversions, Insect hormone biosynthesis, Drug metabolism -cytochrome P450 et.al were activated (Figure S6B). While, arginine and proline metabolism, Proteasome, Phagosome, Oxidative phosphorylation etc. pathway were suppressed (Figure S6B). Mettl51bp mainly affected the physiological activities through activating Phototransduction, Phosphonate metabolism, Glycosylphosphatidylinositol (GPI)-anchor biosynthesis etc. pathways at translational level (Figure S6D). At the same time, it suppressed Peroxisome, amino acid metabolism, Fatty acid degradation, proteasome etc. pathways at the translational level (Figure S6D).

To better understand the correlation between transcription and translation levels, we performed correlation analysis of differentially expressed genes based on fold change of DEGs (Figure 3L). we found 977 genes which had opposite tendency of change at transcriptional vs. translational levels, 24 genes which had same tendency of change at transcriptional vs. translational levels., 240 genes which changed only at translational level and 3309 genes which changed only at transcriptional level. These four gene clusters were analyzed by enrichment analysis (Figure S7A-H). The expression of important clock gene Clk, tim etc. were significantly changed at both transcriptional and translational levels (Figure 3L).

To assess net translation alterations and exclude the effects of transcription changes, we analyzed translation efficiency (TE). We found a change of TE in Mettl51bp compared to w1118 (Figure 3M, Table S1). We obtained 1204 genes that had significantly different TE. We performed GO and KEGG enrichment analysis for these genes. GO enrichment indicated that these genes mainly enriched in amino acid metabolic process and small molecule biosynthetic process etc. (Figure 3N). The KEGG enrichment analysis showed that the main pathway are Glycine, serine and threonine metabolism, one carbon pool by folate, etc. (Figure 3O, P).

Ribo-seq revealed the Mettl51bp led to changes of some global translation features

Ribo-seq revealed the global translation features in two groups, w1118 and Mettl51bp, respectively. The correlation coefficient of TE between the three biological replicates of two groups varied from 0.87 to 0.93, indicating a high correlation between the samples of the same genotype (Figure S8A-F). We found that the length of ribosome-protected RNA fragments (RPFs) was approximately 28nt (Figure S10A). The RPFs exhibited a significant 3-nt periodicity. Metagene analysis of individual 28nt reads revealed the distribution of RPFs across the gene locus in w1118 and Mettl51bp, respectively (Figure S8, S9). The starting point of translation is 12nt upstream of start-codon and gradually disappears 15nt from stop-codon (Figure S8G-L). RPFs on the metagene plot distribution around the translation start and translation stop site (Figure S9A-F), different coding frames on CDS, 3’-UTR and 5’-UTR, respectively in two groups of samples all showed the periodicity (Figure S9G-L). As expected, all plots show an enrichment of P-sites in the first frame on the coding sequence but not the UTRs, in accord with ribosome protected fragments from protein coding mRNAs (Figure S9G-L).

Analysis of the proportion of the different types of open reading frame (ORF) indicated an alteration of the percentage of some types of ORFs. Compared to w1118, we found that Mettl51bp have more overlapped dORFs and less non-overlapped dORFs (Figure S10B). Both of the two groups showed the shorter length of translated uORF compared to untranslated uORF (P<0.05) (Figure S10C). Mettl51bp had less dORF and uORF read count (Figure S10D-E). According to the translation potential of uORF, the translated uORF and untranslated uORF can be divided from different sample. Motif analysis was performed on untranslated uORF and translated uORF respectively (Figure S10F-G).

Mutation of Mettl5 altered the codon preference. When comparing the codon occupancy (A-site) correlation between two groups, we find two codons (GAC and GAU) were preferred by Mettl51bp and UCC were preferred by w1118(Figure S10H). The cumulative frequency of these three codons also demonstrated the same trend (Figure S10I-K). The Asp codon GAC and GAU, so we compared the amino acid occupancy of Asp. The result indicated that the frequency of Asp in w1118 was significantly enriched in translation (Figure S10L).

Metagene plot comparing RPFs between Mettl51bp and w1118 revealed the differential translation. The CDS and region around it was divided into 100 equal bins, and the average RPF density was calculated for each bin. The plot visually depicting the differences in ribosome occupancy patterns between Mettl51bp and w1118along the CDS (Figure S10M), around the translation start site (Figure S10N) and around the translation end site (Figure S10O). The plot highlights the divergent ribosome occupancy patterns between Mettl51bp and w1118 along the CDS and near the translation start site, indicated potential variations in translation dynamics and disparities in translation initiation (Figure S10M, O).

Clock gene regulatory loop regulating circadian rhythm was affected by Mettl51bp

We found that multiple clock genes had significant changes in both transcriptional and translational levels. Cry and Clock was significantly upregulated at both transcriptional and translational levels (Figure 4A-B). Tim, per, vri and pdp1 were significantly downregulated at both transcriptional and translational levels (Figure 4A-B). More important, per, vri, and pdp1 were significantly downregulated in translational efficiency (Figure 4C). These genes are components of the clock gene regulatory network which controls the circadian rhythm. The clock neuron morphology was not affected by the mutant (Figure S11).

Clock genes expression mediated the sleep phenotype caused by Mettl5 mutation.

(A-C) Fold changes in clock genes with significant expression level differences between w1118and Mettl51bp were observed in RNA-seq, Ribo-seq, and translation efficiency analyses. (D-E) PER protein immunofluorescence staining in small LNvs (ventral lateral neurons) and its immunofluorescence intensity statistics. (F-G) Western blot analysis of PER protein and Tubulin protein, along with their grayscale quantification. (H) Sleep curves throughout the day for Mettl51bp, CLKJRK, double mutant and control flies. (I) Total sleep for Mettl51bp, CLKJRK, double mutant and control flies. (J) Percentage of awake time in Mettl51bpflies, partially rescued by double mutant flies. (K) Sleep curve throughout the day for Mettl51bp, per01, double mutant and control flies. (L) Total sleep for Mettl51bp, per01, double mutant and control flies. (M) Percentage of awake time in Mettl51bp flies, partially rescued by double mutant flies. (N) Fold changes in proteasome subunits with significant expression level differences between w1118 and Mettl51bp were observed in RNA-seq, Ribo-seq, and translation efficiency analyses. For statistical significance, * stands for p<0.05, ** stands for p<0.01, *** stands for p<0.001, ns stands for not significant.

According to the relationship between these genes in the clock regulatory circuit, if Per was downstream of Clock in mediating the function of Mettl5, we should have seen the change of both genes to the same direction. However, we detected changes to opposite for Clock and Per. Therefore, we propose that Per was upstream of Clock in mediating the function of Mettl5. Then, we examined the protein level of Per. Surprisingly, we found that the protein level of Per was upregulated in the mutant detected by both immunostaining and western blot (Figure 4D-G). This is confirmed by the circadian rhythm phenotype showed by Mettl5 mutant, which showed a significant increased period length (Table 1). An increased period length was reported in previous case of inactive kinase activity which results in a stabilized Period (Philpott et al., 2023). Consistently, genetic evidence indicated that both of Clock and Per genes were downstream of Mettl5, because the double mutant showed the phenotype of clock genes (Figure 4H-M, Table 1). This result indicated that other mechanism regulating Per protein was affected by Mettl51bp.

Then we ask what’s the factors mediated Per protein level changes in Mettl5 mutant. By examining the RNA-seq, Ribo-seq data and translation efficiency, we found that multiple components of the proteasome pathways were significantly downregulated in Mettl51bp (Figure 3E, I, 4N). Based on the above results, we proposed a working model in which Mettl5 directly regulates proteasome, Per and Clock in transcription and translation. The proteasome regulates Per at post translational level. And in turn Per inhibits Clock at protein level. As a result, in Mettl51bp, the down regulation of proteasome results in the upregulation of Per protein and as a result up regulation of Clock protein activity and its downstream gene transcription (Figure 5).

A working model illustrating the role of Mettl5 in Drosophila sleep was presented.

The axon complexity was affected by Mettl51bp

Previous results have indicated a close correlation between sleep need and synaptic complexity (Bushey et al., 2011). Moreover, it has been observed that synaptogenesis stimulates proteasome activity in axons (Costa et al., 2019). Considering the loss of sleep rebound after deprivation and the expression level changes of proteasome subunits resulting from Mettl51bp, we were interested in investigating whether synaptic complexity is affected.

To quantify synaptic complexity, we employed a method previously reported (Bushey et al., 2011). Syt-GFP, a protein product that colocalizes with native synaptic vesicles, was used to visualize changes in presynaptic morphology. As controls, UAS-Fmr1 and Fmr1 mutation were utilized, resulting in a decrease and increase of syt-eGFP, respectively (Figure 6A-C). Interestingly, we found that the Mettl51bp led to an increase in syt-eGFP staining in presynaptic terminals (Figure 6D).

The axon complexity was found to be affected by Mettl51bp

(A-D) Examples of small LNvs axonal terminals labeled by syt-eGFP (green). A rectangular box indicates the location where LNvs axons exit the posterior optic tract and the first axonal bifurcation to the tip of the terminal region, whose volume was measured. (E) Mean volume measurements of small LNvs axonal terminals labeled by syt-eGFP.

Discussion

In this study, we found that Mettl5, a previously identified rRNA methylase, regulates sleep through its RNA methylation activity. RNA-seq and Ribo-seq of the Mettl51bp identified downstream genes of Mettl5 at transcriptional and translational levels. Further study identified Mettl5 can affect clock gene network and proteasome pathway which mediated Mettl5 regulation of sleep. By elucidated the molecular mechanism for Mettl5 function on sleep, this study highlighted an integrative role of protein synthesis and degradation in sleep regulation. The Ribo-seq analysis in this study indicated that global features of translation, such as uORF translation efficiency and codon choices, were affected in the Mettl51bp, suggesting a role of rRNA methylation in these processes.

The ortholog of Mettl5 in humans is METTL5, mutation of which causes intellectual disability (ID) (Richard et al., 2019). This study expanded our understanding of the molecular mechanism of Mettl5 function, which potentially provides new insights for the potential treatment of ID. Previous clinical studies found that the ID patients have sleep problems, including sleep duration shortness. This symptom is consistent with what we found for the Mettl51bp. The mechanism for this phenotype found in this study is potentially important. More studies need to be done in vertebrates to verify if this mechanism is conserved for sleep disorders.

The RNA-seq and Ribo-seq of Mettl51bp revealed the gene network that is potentially directed by Mettl5. The direct or immediate effect of Mettl5 may lie in the regulation of translation. Indeed, we found that the codon occupancy was significantly changed for GAC, GAU, UCC. Moreover, the occupancy on dORF were significantly downregulated in the Mettl51bp. The translated uORF were significantly downregulated by the Mettl51bp. These may lead to fluctuations in certain protein production, which may eventually lead to altered downstream gene expression. This result indicated that small perturbations in translation process have a significant impact on downstream events, highlighting the importance of maintaining the accuracy of the translation process.

We observed a significant decrease in total RNA and 18S rRNA methylation levels in individuals with the heterozygous allele of Mettl5. Interestingly, this finding differs slightly from studies conducted in mice, where heterozygous knockouts still exhibited normal levels of 18S rRNA m6A methylation (Sepich-Poore et al., 2022). It may attribute to potential species-specific regulatory differences between Drosophila and mice. Another plausible explanation may lie in the distinct nature of the knockdown compared to knockouts. Notably, prior research has highlighted compensation effects arising from complete knockouts (Teng et al., 2013; Rossi et al., 2015; Vu et al., 2015; Ma et al., 2019; El-Brolosy et al., 2019).

This study revealed a new aspect on the connection of clock genes and proteasome. The clock regulation on proteasome was important in several different contexts. Previous studies showed that the circadian clock rhythmically controls the daily expression of multiple components of the proteasome, which responds to dietary restriction in Drosophila fat body (Hwangbo et al., 2023). In this study, we found Mettl5 can regulate clock protein synthesis and degradation in clock neurons via affecting proteasome function. The role played by Mettl5 in this process may be through affecting the methylation of ribosomes in clock neurons, affecting the proteasome degradation pathway (Costa et al., 2019), and ultimately leading to changes in clock proteins in clock neurons.

The function of Mettl5 highlights an integrative regulation of rRNA methylation and the proteasome. This provides an opportunity to coordinate protein synthesis and protein degradation, which would be a feasible way to keep balance for the total protein level. Interestingly, we observed specificity in the downstream affected genes both at transcriptional and translational levels. For the proteasome subunits, we also detected level changes for some of them, indicating a specificity of Mettl5 function. The specificity for the proteasome subunits changes could be attributed to gene specificity of Mettl5 on translation. Another possibility is that there are other layers of regulation on protein degradation pathway by Mettl5 protein interaction or other mechanisms. These factors may form a regulatory network to coordinate with each other. It is an intriguing question how the downstream specificity is achieved. More studies are needed to elucidate these mechanisms.

This result indicates the complex relationship between clock genes and sleep homeostasis regulation. The data in this study are consistent with previous reports on the involvement of clock genes in the regulation of sleep homeostasis. It has been reported that loss of function alleles of Per and Clock exhibit a more pronounced sleep rebound after sleep deprivation (Shaw et al., 2002). Conversely, the Mettl5 allele, which results in increased PER protein, leads to reduced sleep rebound. Furthermore, data in this study suggested that the regulatory mechanism of Clock has tissue specificity. According to the expression patterns of Mettl5 based on Mettl5-Gal4, it does not express in clock neurons. Instead, it is expressed in Mettl5-Gal4 positive cells, which are composed of a part of neurons and glia. Clock is expressed not only in clock neurons of the Drosophila brain (Patop et al., 2023). The combinational expression of these proteins constitutes the specificity of clock upstream regulators. Our study indicates that these Mettl5 expressing neurons and glia are important in sleep regulation. It would be interesting to find out the Mettl5 dependent sleep homeostasis regulatory circuits.

Evidence indicates that sleep functions in development, metabolism and neuronal plasticity (Anafi et al., 2019). Sleep exerts effects on neuronal plasticity by modifying synapses. The synaptic homeostasis theory proposes that sleep has a role in downscale synaptic strength (Tononi and Cirelli, 2006). Indeed, synapse markers progressively decrease during sleep in both mammals and Drosophila brains (López-Varea et al., 2021; Vyazovskiy et al., 2008; Gilestro et al., 2009; Liu et al., 2010) More importantly, evidence indicates that sleep need and protein levels in synapse are tightly linked. Studies of the presynaptic active zone have shown that synaptic plasticity regulates sleep homeostasis (Huang et al., 2020). Genome-wide proteomic studies of synapse in mouse brain indicate that the synaptic proteins peak around dusk and dawn (Noya et al., 2019; Brüning et al., 2019). Sleep deprivation experiments have demonstrated that the sleep drive is significantly dependent on the cycling of proteins and phosphoproteins in synapses, in contrast to mRNAs (Noya et al., 2019; Brüning et al., 2019). These results suggest that the gene expression regulation at the protein level is crucial for sleep. Consistently, we detected changes in the quantity of axons in Mettl51bp, suggesting a possible regulation of neural circuits by Mettl5.

This study revealed gene-specific functions of ribosome components. Previous study suggested a context dependent component and function of ribosomes (Simsek et al., 2017). We observed changes in the translational efficiency of genes in the Mettl5 mutant. One possible mechanism is that rRNA modifications influenced by Mettl5 affect the binding affinity of ribosomes to specific RNA sequences. Alternatively, the rRNA modifications influenced by Mettl5 may be responsible for certain ribosome types that impact the translation of this gene profile. Further evidence is needed to elucidate these possibilities.

Material and methods

Drosophila Strains

Fly stocks used in this study were maintained under standard culture conditions. We used the w1118 as the control strain. The following flies were obtained from Bloomington Stock Center: w1118 (Bl: 5905), Mettl5-Gal4 (Bl: 19514), Repo-Gal4 (Bl: 7415), nSyb-Gal4 (Bl: 51941), pdf-Gal4 (Bl: 41286), ClkJRK (Bl: 80927), per01(Bl: 80917), Fmr1Δ50M(Bl: 6930), UAS-Trmt112RNAi (VDRC: 101515), UAS-DenMark,UAS-syt.eGFP (Bl: 33064), UAS-GFPstinger (from Yi Rao’s lab), UAS-Fmr1 (Bl: 6931), UAS-Mettl5 (FlyORF: F000760) are from Fly ORF collection. UAS-Mettl5m-3HA (generated in this study). Mettl51bpand Mettl59bp were generated by the CRISPR/Cas system.

Sleep behavior assays

Sleep assays were performed in an incubator at a temperature of 25 ± 1 ℃ and a relative humidity of 60% ± 5%. Lights were turned on at ZT0 (06:30 am) and off at ZT12 (18:30 pm). The locomotor activity was recorded using Drosophila Activity Monitoring System (Trikinetics. Waltham, MA). Flies were acclimated to the experimental conditions for 2 days before data collection. Locomotor activity data were collected at 1-minute intervals for 3 days, and analyzed by pysolo as described previously (Gilestro and Cirelli, 2009). Sleep was defined as periods of 5 min or more of behavioral immobility.

Sleep deprivation is applied by mechanical stimuli using a timer-controlled rotating shaker. The intensity of stimuli was 1500 rpm once every minute (2 s/1 min) (Shimizu et al., 2008). After acclimating to the experimental conditions for 2 days, the flies were sleep deprived for 12 h (ZT12-ZT24). Mechanical sleep deprivation was performed using the SNAP method to keep flies awake for 12 h overnight (Shaw et al., 2002). Flies were sleep deprived from ZT12 (beginning of the dark phase) to ZT0 (beginning of the light phase) by a vortex at the lower intensity setting for 3 s every minute. Sleep lost and recovery were calculated for each fly by using the 24-h period preceding deprivation as the baseline. Sleep recovery values were calculated by subtracting 6 h sleep amount during night (ZT12-ZT18) after deprivation from the sleep amount of the corresponding time of the previous day (baseline sleep amount).

Sleep arousal is measured at ZT14. Flies are given a uniform, gentle mechanical stimulus, and any activity within the next five minutes is recorded. If no activity is observed during this period, the fly is considered not to have awakened in response to the stimulation. The number of flies that awaken due to the stimulus is counted, and their percentage is calculated.

Statistical analysis

Statistical analysis was carried out in GraphPad Prism 5. The statistical tests for each experiment are shown in the figures. The significance of difference for sleep parameters was analyzed using Nonparametric test, including two-tailed Mann-Whitney test or one-way ANOVA with Tukey’s multiple comparison test. Unpaired Student’s t-test analyses were carried out for qPCR data and Log-rank (Mantel-Cox) test analyses were carried out for lifespan data using GraphPad Prism software. For all statistical tests, p < 0.05 was considered significant.

Quantitative PCR

Total RNA was extracted from cells and tissues using the TRNzol Universal Reagent (Tiangen #DP4-02). RNA reverse transcription was performed by using PrimeScript RT reagent Kit with gDNA Eraser (TAKARA #RR047A). Quantitative PCR was carried out using SuperReal PreMix Plus (SYBR Green) (Tiangen #DP4-02). RP49 were used to normalize results between groups. Three independent biological replicates were performed for each experiment. The primers used in this experiment are RP49-F: CGGTTACGGATCGAACAAGC; RP49-R: CTTGCGCTTCTTGGAGGAGA; Mettl5-F: CAACACGGTCGTACATTCAG; Mettl5-R: AGTCCACCTC AATATCCTTCG.

Imaging and analysis of axon volume

To quantify the axon volume of small LNvs, the plugin Object 3D counter in ImageJ was used to quantify number of pixels from the raw images. The images were processed according to a threshold in which an intact axon can be observed in controls. Axon span from the first axonal bifurcation to the tip of the terminal which was labeled by white rectangular box in Figure 8 was quantified.

Sample collection and library construction for Ribo-seq and RNAseq

For Drosophila samples, the w1118 and Mettl51bp at ZT15 were transferred to centrifuge tubes and quick-frozen by immersion in liquid nitrogen. The fly heads were separated and collected after quick-frozen with liquid nitrogen. Three biological replicates were collected for w1118and Mettl51bp, respectively. Half of each sample was used for RNA-seq and the other half for Ribo-seq.

For Ribo-seq library construction, samples were treated with specific lysis buffer with cycloheximide (50mg/mL) to acquire the lysate (Novogen, China). To digest the RNA other than ribosome-protected fragments (RPF), fly tissue lysate was treated with unspecific endoribonuclease, RNase I. Isolation of monosomes was performed by size-exclusion chromatography using MicroSpin S-400 HR columns. The RNA samples were then treated with rRNA depletion kit to remove as much rRNA contamination as possible before performing PAGE purification of the relatively short (20∼38 nt) RPFs. Following PAGE purification, both ends of the RPF were phosphorylated and ligated with 5’ and 3’ adapters, respectively. Then the fragments were reversely transcribed to the cDNAs, amplified by PCR, and subjected to library construction after quality test. Ribo-seq libraries were sequenced using PE150 (Illumina).

For RNA-seq library construction, total RNA was extracted using a standard TRIzol RNA extraction protocol. The cDNA library was built and sequenced using illumina PE 150 (Illumina). The RNAseq analysis was performed by Novogen.

Ribo-seq and RNA-seq data were analyzed as follows. Raw data (raw reads) of FASTQ format were first processed. Clean reads were obtained by removing reads containing adapters, reads containing N bases and low-quality reads from the raw data. At the same time, Q20, Q30 and GC content of the clean data were calculated.

RNA-seq analysis

Index of the Drosophila melanogaster reference genome (dm6) was built using Hisat2 v2.0.5 (Kim et al., 2019) and paired-end clean reads were aligned to the Drosophila melanogaster reference genome (dm6) using Hisat2 v2.0.5. The mapped reads of each sample were assembled by StringTie (v1.3.3b) (Pertea et al., 2015) in a reference-based approach. FeatureCounts (v1.5.0) (Liao et al., 2014) was used to count the reads numbers mapped to each gene. Differential expression analysis of w1118 and Mettl51bp was performed using the DESeq2 R package (1.20.0) (Pertea et al., 2015). Genes with an adjusted P-value<0.05 and |log2fc| ≥1 were assigned as significant differentially expressed gene.

Ribo-seq analysis

First, the clean reads were aligned to a FASTA file containing rRNA, tRNA, snoRNA, and snRNA from Flybase Release 6.13 annotations using Bowtie (Liao et al., 2014) with a maximum of two mismatches (-v 2) by default. Successful alignments were discarded, and all unaligned reads were collected. Unaligned clean reads (ribosome-protected fragments, RPF) were then mapped to the FlyBase D. melanogaster reference genome (Release 6.13) using STAR (v2.7.3a) (Dobin et al., 2013) with default options. The unique genome-mapped reads are then mapped against protein coding transcripts using Bowtie (v1.2.2) (Langmead et al., 2009) with parameters “-a -v 2” by default. The featureCounts program in Subread package v1.6.3 is used to count the number of RPFS uniquely mapped to CDS regions based on fly genome mapping file (-t CDS -g gene_id), which were then normalized as RPF Per Kilobase per Million mapped RPFs (RPKM). DESeq2 R package (1.14.1) (Love et al., 2014) was used for differential expression analysis. The threshold for significantly differential expression is |log2 (Fold Change) | ≥ 0.265 & padj< 0.05. Differential TE analysis was performed using RiboDiff. Coding frame distribution and 3-nt periodicity analyses for Ribo-seq quality evaluation were performed using R package riboWaltz (v1.1.0) (Lauria et al., 2018). The Ribocode package (Xiao et al., 2018) was used to analyze the P-site and the motif of translated/untranslated uORFs. Gene Ontology (GO) and KEGG pathway analysis, Gene Set Enrichment Analysis (GSEA) analysis was performed using the clusterProfiler package (v4.5.2.002) (Wu et al., 2021). Volcano plots, veen plots a were plotted with the ggplot2 package. SimplifyEnrichment package (Gu and Hübschmann, 2022) was used to simply the GO enrichment results.

Immunofluorescence experiments

Adult flies that were 7-15 days old (unless otherwise noted) were anaesthetized with CO2 and dissected in 0.03% PBST (1 × PBS with 0.03% Triton X-100; Sigma, T9284) on ice. After a 55 min fixation of samples in 2% paraformaldehyde (PFA) at room temperature (RT), samples were washed four times for 15 min in 0.03% PBST at RT, blocked in 10% Normal Goat Serum (NGS; diluted with 1 × PBS with 2% Triton) overnight at 4°C and incubated with primary antibodies for 24 h at 4°C. The samples were washed again four times with 1 × PBS with 1% Triton for 15 min at RT and incubated with secondary antibodies overnight at 4 °C. And then the samples were washed with 1 × PBS with 1% Triton again 4 times for 15 min at RT and mounted on a slide using anti-fading Mounting medium (with DAPI; Solarbio, S2110). The primary antibodies we used were anti-Elav rat (DSHB, 7E8A10, 1:200) and anti-Repo mouse (Developmental Studies Hybridoma Bank, 8D12; 1:200). The secondary antibodies were Alexa FluorTM 568 (Thermo Fisher Scientific, A11004; 1:200) and Alexa FluorTM 647 (Thermo Fisher Scientific, A21247; 1:200). The primary and secondary antibodies were diluted in dilution buffer (1.25% PBST, 1% NGS). Images were taken using confocal microscopy (Leica SP8) with LAX software with auto Z brightness correction to generate a homogeneous signal where it seemed necessary and were formatted using Adobe Photoshop CS6. Figures were generated using Adobe IIIustrator 2020. For the immunofluorescence experiment of PDF, the method is similar with the discription above. All immunofluorescence experiments were done with at least 3 repeats, each repeat with a sample size of more than 10 individual flies. ImageJ was used to analyze immunofluorescence brightness.

Western Blotting

Total protein was extracted from approximately 30 fly heads using RIPA buffer (150 mM NaCl, 1.0% NP-40, 0.5% sodium deoxycholate, 0.1% SDS, 50 mM Tris-HCl, pH 8.0). Protease inhibitor cocktail (CW2200S) and phosphatase inhibitor cocktail (CW2383S) were added according to the manufacturer’s instructions. An equal volume of 2x SDS loading buffer was added to the samples, followed by heating in a water bath at 100 °C for 5 minutes, and then cooling on ice for several minutes.

Primary antibodies (Rabbit anti-PER, Gift from Dr. Jeffrey Price’s laboratory at University of Missouri at Kansas City, U.S.) were diluted 1:5000 and incubated with the membranes overnight at 4°C, while HRP-labeled secondary antibodies (HRP Goat Anti-Rabbit IgG (H + L) (ABclonal, AS014)) were diluted 1:1500 and incubated for 4 hours at room temperature. Signals were detected using ECL (ABclonal, RM00021P) and visualized with the Amersham ImageQuant 800 (GE Healthcare, Sweden). Signal intensity was quantified using ImageJ software. Three biological replicates were conducted for this experiment.

LC-MS/MS analysis of m6A levels

Total RNA was extracted from cells and tissues using TRNzol Universal Reagent (Tiangen #DP4-02). 18S rRNA was obtained through polyacrylamide gel electrophoresis and gel recovery of total RNA. Ribonucleoside standards included rA and N6 mA. The fluid phase used in this experiment was a mixture of methanol and ddH2O. For each biological replicate, 1 μg of total RNA or 18S rRNA was degraded to single nucleotides using Nucleoside Digestion Mix (NEB #M0649), followed by the addition of a 4-fold volume of methanol and precipitation at −20°C for 2 hours to remove proteins. Multiple reaction monitoring transitions used in this study were 268.10275 → 136.0621 (rA), 282.11835→150.0774 (N6 mA).

Lifespan and Negtive geotaxis RING assay

Mated males were collected and maintained at a density of 20 flies per vial. The flies were kept at a temperature of 25 °C with 60% humidity and a 12-hour light/12-hour dark cycle. Every other day, the flies were transferred to new vials, and their survival rate was recorded. Statistical analysis and survival curve plotting were performed using GraphPad Prism 5.0. Each experiment was conducted with three biological replicates, using a total of 300 flies for lifespan assays.

For climbing assays, flies were collected under CO2 anesthesia and allowed to recover overnight. Typically, 200 flies (distributed among 10 vials, each containing 20 flies and marked at 90 mm from the bottom) were used for each genotype. To initiate the climbing assay, vials were tapped sharply on the table three times to elicit negative geotaxis responses. A 15-minute rest period was observed between each trial. Three trials were conducted for each group, and five groups were used for each genotype. The number of flies attempting to reach the mark within 10 seconds was counted for each trial. Videos were recorded during the assay, and the positions of the flies were captured. The camera and vials were separated by a 20 cm distance, and a white open-faced box placed behind the setup provided uniform lighting.

eLife Assessment

In this manuscript, the authors present useful findings demonstrating that the RNA modification enzyme Mettl5 regulates sleep in Drosophila. Through transcriptome- and proteome-wide analyses, the authors identified downstream targets affected in heterozygous mutants and proposed that Mettl5 regulates the translation and degradation of clock genes to maintain normal sleep function. However, the mechanisms by which Mettl5 achieves these functions, and whether they are direct or indirect, remain incomplete and would benefit from further analysis.

Reviewer #1 (Public review):

Summary:

Here the authors attempted to test whether the function of Mettl5 in sleep regulation was conserved in drosophila, and if so, by which molecular mechanisms. To do so they performed sleep analysis, as well as RNA-seq and ribo-seq in order to identify the downstream targets. They found that the loss of one copy of Mettl5 affects sleep and that its catalytic activity is important for this function. Transcriptional and proteomic analyses show that multiple pathways were altered, including the clock signaling pathway and the proteasome. Based on these changes the authors propose that Mettl5 modulate sleep through regulation of the clock genes, both at the level of their production and degradation.

Strengths:

The phenotypical consequence of the loss of one copy of Mettl5 on sleep function is clear and well-documented.

Weaknesses:

The imaging and molecular parts are less convincing.

- The colocalization of Mettl5 with glial and neuronal cells is not very clear

- The section on gene ontology analysis is long and confusing

- Among all the pathways affected the focus on proteosome sounds like cherry picking. And there is no experiment demonstrating its impact in the Mettl5 phenotype

- The ribo seq shows some changes at the level of translation efficiency but there is no connection with the Mettl5 phenotypes. In other words, how the increased usage of some codons impact clock signalling. Are the genes enriched for these codons?

- A few papers already demonstrated the role of Mettl5 in translation, even at the structural level (Rong et al, Cell reports 2020) and this was not commented by the authors. In Peng et al, 2022 the authors show that the m6A bridges the 18S rRNA with RPL24. Is this conserved in Drosophila?

- The text will require strong editing and the authors should check and review extensively for improvements to the use of English.

Conclusion

Despite the effort to identify the underlying molecular defects following the loss of Mettl5 the authors felt short in doing so. Some of the results are over-interpreted and more experiments will be needed to understand how Mettl5 controls the translation of its targets. References to previous works was poorly commented.

Reviewer #2 (Public review):

Summary:

The authors define the m6A methyltransferase Mettl5 as a novel sleep-regulatory gene that contributes to specific aspects of Drosophila sleep behaviors (i.e., sleep drive and arousal at early night; sleep homeostasis) and propose the possible implication of Mettl5-dependent clocks in this process. The model was primarily based on the assessment of sleep changes upon genetic/transgenic manipulations of Mettl5 expression (including CRISPR-deletion allele); differentially expressed genes between wild-type vs. Mettl5 mutant; and interaction effects of Mettl5 and clock genes on sleep. These findings exemplify how a subclass of m6A modifications (i.e., Mettl5-dependent m6A) and possible epi-transcriptomic control of gene expression could impact animal behaviors.

Strengths:

Comprehensive DEG analyses between control and Mettl5 mutant flies reveal the landscape of Mettl5-dependent gene regulation at both transcriptome and translatome levels. The molecular/genetic features underlying Mettl5-dependent gene expression may provide important clues to molecular substrates for circadian clocks, sleep, and other physiology relevant to Mettl5 function in Drosophila.

Weaknesses:

While these findings indicate the potential implication of Mettl5-dependent gene regulation in circadian clocks and sleep, several key data require substantial improvement and rigor of experimental design and data interpretation for fair conclusions. Weaknesses of this study and possible complications in the original observations include but are not limited to:

(1) Genetic backgrounds in Mettl5 mutants: the heterozygosity of Mettl5 deletion causes sleep suppression at early night and long-period rhythms in circadian behaviors. The transgenic rescue using Gal4/UAS may support the specificity of the Mettl5 effects on sleep. However, it does not necessarily exclude the possibility that the Mettl5 deletion stocks somehow acquired long-period mutation allelic to other clock genes. Additional genetic/transgenic models of Mettl5 (e.g., homozygous or trans-heterozygous mutants of independent Mettl5 alleles; Mettl5 RNAi etc.) can address the background issue and determine 1) whether sleep suppression tightly correlates with long-period rhythms in Mettl5 mutants; and 2) whether Mettl5 effects are actually mapped to circadian pacemaker neurons (e.g., PDF- or tim-positive neurons) to affect circadian behaviors, clock gene expression, and synaptic plasticity in a cell-autonomous manner and thereby regulate sleep. Unfortunately, most experiments in the current study rely on a single genetic model (i.e., Mettl5 heterozygous mutant).

(2) Gene expression and synaptic plasticity: gene expression profiles and the synaptic plasticity should be assessed by multiple time-point analyses since 1) they display high-amplitude oscillations over the 24-h window and 2) any phase-delaying mutation (e.g., Mettl5 deletion) could significantly affect their circadian changes. The current study performed a single time-point assessment of circadian clock/synaptic gene expression, misleading the conclusion for Mettl5 effects. Considering long-period rhythms in Mettl5 mutant clocks, transcriptome/translatome profiles in Mettl5 cannot distinguish between direct vs. indirect targets of Mettl5 (i.e., gene regulation by the loss of Mettl5-dependent m6A vs. by the delayed circadian phase in Mettl5 mutants).

(3) The text description for gene expression profiling and Mettl5-dependent gene regulation was very detailed, yet there is a huge gap between gene expression profiling and sleep/behavioral analyses. The model in Figure 5 should be better addressed and validated.

Reviewer #3 (Public review):

Xiaoyu Wu and colleagues examined the potential role in sleep of a Drosophila ribosomal RNA methyltransferase, mettl5. Based on sleep defects reported in CRISPR generated mutants, the authors performed both RNA-seq and Ribo-seq analyses of head tissue from mutants and compared to control animals collected at the same time point. While these data were subjected to a thorough analysis, it was difficult to understand the relative direction of differential expression between the two genotypes. In any case, a major conclusion was that the mutant showed altered expression of circadian clock genes, and that the altered expression of the period gene in particular accounted for the sleep defect reported in the mettl5 mutant. As noted above, a strength of this work is its relevance to a human developmental disorder as well as the transcriptomic and ribosomal profiling of the mutant. However, there are numerous weaknesses in the manuscript, most of which stem from misinterpretation of the findings, some methodological approaches, and also a lack of method detail provided. The authors seemed to have missed a major phenotype associated with the mettl5 mutant, which is that it caused a significant increase in period length, which was apparent even in a light: dark cycle. Thus the effect of the mutant on clock gene expression more likely contributed to this phenotype than any associated with changes in sleep behavior.

Author response:

The following is the authors’ response to the original reviews.

Public Reviews:

Reviewer #1 (Public review):

Summary:

Here the authors attempted to test whether the function of Mettl5 in sleep regulation was conserved in drosophila, and if so, by which molecular mechanisms. To do so they performed sleep analysis, as well as RNA-seq and ribo-seq in order to identify the downstream targets. They found that the loss of one copy of Mettl5 affects sleep and that its catalytic activity is important for this function. Transcriptional and proteomic analyses show that multiple pathways were altered, including the clock signaling pathway and the proteasome. Based on these changes the authors propose that Mettl5 modulate sleep through regulation of the clock genes, both at the level of their production and degradation.

Strengths:

The phenotypical consequence of the loss of one copy of Mettl5 on sleep function is clear and well-documented.

Weaknesses:

The imaging and molecular parts are less convincing.

- The colocalization of Mettl5 with glial and neuronal cells is not very clear

We truly appreciate your suggestion. We repeated the staining experiments. To ensure better results, we tried another antibody of ELAV (mouse) and optimized the experimental conditions. This result has been included in the Figure S1 of the revised version.

- The section on gene ontology analysis is long and confusing

The session is revised for clarity. To get a better flow of logic, we deleted the paragraph which describing the details of Figure S6.

- Among all the pathways affected the focus on proteosome sounds like cherry picking. And there is no experiment demonstrating its impact in the Mettl5 phenotype

Thank you for the comments. The changes of period oppositely at transcriptional versus translational levels puzzled us a while until we found the ubiquitin pathway components changes. The regulation of Period protein degradation by ubiquitin-proteasome pathway has been well documented (Grima et al., 2002; Ko et al., 2002; Chiu et al., 2008). In addition, previous reports indicated that N6 methyladenosine (m6A) regulates ubiquitin proteasome pathway in skeletal muscle physiology (Sun et al., 2023). This information has been included in the revised manuscript in the last paragraph under the title: Mettl5 regulates the clock gene regulatory loop.

Indeed, we haven’t found a proper way to manipulate proteasome levels in genetic tests. Proteasome is a large protein complex which is composed of many subunits. Enhancing the its activity by overexpressing its components was not applicable. Moreover, proteasome has important function during many biological processed. Disrupting its function by simply MG132 treatment which we tried results in lots of side effects.

In this study, we also noticed the codon usage alteration caused by mettl5 mutant. Please refer to the answers to the following question for details. Previous reports also found the regulation of mettl5 on translation in other systems (Rong et al, 2020; Peng et al., 2022). Based on these analyses, it is possible that both the regulation on translation and protein degradation contributed the period protein upregulation found in mettl5 mutant. This idea has been included in the Discussion session of the revised manuscript.

References

Sun J, Zhou H, Chen Z, et al. Altered m6A RNA methylation governs denervation-induced muscle atrophy by regulating ubiquitin proteasome pathway. J Transl Med. 2023;21(1):845. Published 2023 Nov 23. doi:10.1186/s12967-023-04694-3

Grima, B. et al. The F-box protein slimb controls the levels of clock proteins period and timeless. Nature 420, 178–182 (2002).

Ko, H. W., Jiang, J. & Edery, I. Role for Slimb in the degradation of Drosophila period protein phosphorylated by doubletime. Nature 420, 673–678 (2002).

Chiu, J. C., Vanselow, J. T., Kramer, A. & Edery, I. The phosphooccupancy of an atypical SLIMB-binding site on PERIOD that is phosphorylated by DOUBLETIME controls the pace of the clock. Genes Dev. 22, 1758–1772 (2008).

- The ribo seq shows some changes at the level of translation efficiency but there is no connection with the Mettl5 phenotypes. In other words, how the increased usage of some codons impact clock signalling. Are the genes enriched for these codons?

Thank you for raising this point. In our analysis, we observed an increased usage of the codons for Asp in the Mettl5 mutant. Prior work has reported a possible connection between codon usage and per protein activity. In the report, a per version with optimized codon cannot rescue circadian rhythmicity caused by per mutant, in contrast to WT version (Fu J et al. 2016). Further study indicated that dPER protein levels were also elevated in the mutant flies, suggesting a role for codon optimization in enhancing dPER expression (Figure 2B in Fu J et al. 2016). Consistent with this, we analyzed the region of codon optimization in Fu J et al. 2016. The result indicated that that GAC has a relatively high usage rate in these regions (indicated in the following two Author response image charts by the red arrow), suggesting that the Mettl5 mutation may influence per protein accumulation through altered GAC usage. Further experiments are needed to confirm this possibility. We included these details in the second last paragraph of the Discussion session.

15-21

SDSAYSN

43-316

SSGSSGYGGKPSTQASSSDMIIKRNKEKSRKKKKPKCIALATATTVSLEGTEESPLPANGGCEKVLQELQDTQQLGEPLVVTETQLSEQLLETEQNEDQNKSEQLAQFPLPTPIVTTLSPGIGPGHDCVGGASGGAVAGGCSVVGAGTDKTSELIPGKLESAGTKPSQERPKEESFCCVISMHDGIVLYTTPSISDVLGFPRDMWLGRSFIDFVHHKDRATFASQITTGIPIAESRGCMPKDARSTFCVMLRRYRGLNSGGFGVIGRAVNYEPF

Fu J, Murphy KA, Zhou M, Li YH, Lam VH, Tabuloc CA, Chiu JC, Liu Y. Codon usage affects the structure and function of the Drosophila circadian clock protein PERIOD. Genes Dev. 2016 Aug 1;30(15):1761-75.

- A few papers already demonstrated the role of Mettl5 in translation, even at the structural level (Rong et al, Cell reports 2020) and this was not commented by the authors. In Peng et al, 2022 the authors show that the m6A bridges the 18S rRNA with RPL24. Is this conserved in Drosophila?

Thanks for the reminder. We discussed and cited these papers in the revised version.

Rong B, Zhang Q, Wan J, et al. Ribosome 18S m6A Methyltransferase METTL5 Promotes Translation Initiation and Breast Cancer Cell Growth. Cell Rep. 2020;33(12):108544. doi:10.1016/j.celrep.2020.108544

Peng H, Chen B, Wei W, et al. N6-methyladenosine (m6A) in 18S rRNA promotes fatty acid metabolism and oncogenic transformation. Nat Metab. 2022;4(8):1041-1054. doi:10.1038/s42255-022-00622-9

- The text will require strong editing and the authors should check and review extensively for improvements to the use of English.

Thanks. The text of the paper are thoroughly revised.

Conclusion

Despite the effort to identify the underlying molecular defects following the loss of Mettl5 the authors felt short in doing so. Some of the results are over-interpreted and more experiments will be needed to understand how Mettl5 controls the translation of its targets. References to previous works was poorly commented.

Thanks for your suggestion. We have incorporated the references mentioned above. However, our efforts have thus far fallen short of elucidating a precise picture of METTL5's functional mechanism. To address this, the limitations of the current study have been discussed more thoroughly in the revised main text.

Reviewer #2 (Public review):

Summary:

The authors define the m6A methyltransferase Mettl5 as a novel sleep-regulatory gene that contributes to specific aspects of Drosophila sleep behaviors (i.e., sleep drive and arousal at early night; sleep homeostasis) and propose the possible implication of Mettl5-dependent clocks in this process. The model was primarily based on the assessment of sleep changes upon genetic/transgenic manipulations of Mettl5 expression (including CRISPR-deletion allele); differentially expressed genes between wild-type vs. Mettl5 mutant; and interaction effects of Mettl5 and clock genes on sleep. These findings exemplify how a subclass of m6A modifications (i.e., Mettl5-dependent m6A) and possible epi-transcriptomic control of gene expression could impact animal behaviors.

Strengths:

Comprehensive DEG analyses between control and Mettl5 mutant flies reveal the landscape of Mettl5-dependent gene regulation at both transcriptome and translatome levels. The molecular/genetic features underlying Mettl5-dependent gene expression may provide important clues to molecular substrates for circadian clocks, sleep, and other physiology relevant to Mettl5 function in Drosophila.

Weaknesses:

While these findings indicate the potential implication of Mettl5-dependent gene regulation in circadian clocks and sleep, several key data require substantial improvement and rigor of experimental design and data interpretation for fair conclusions. Weaknesses of this study and possible complications in the original observations include but are not limited to:

(1) Genetic backgrounds in Mettl5 mutants: the heterozygosity of Mettl5 deletion causes sleep suppression at early night and long-period rhythms in circadian behaviors. The transgenic rescue using Gal4/UAS may support the specificity of the Mettl5 effects on sleep. However, it does not necessarily exclude the possibility that the Mettl5 deletion stocks somehow acquired long-period mutation allelic to other clock genes. Additional genetic/transgenic models of Mettl5 (e.g., homozygous or trans-heterozygous mutants of independent Mettl5 alleles; Mettl5 RNAi etc.) can address the background issue and determine 1) whether sleep suppression tightly correlates with long-period rhythms in Mettl5 mutants; and 2) whether Mettl5 effects are actually mapped to circadian pacemaker neurons (e.g., PDF- or tim-positive neurons) to affect circadian behaviors, clock gene expression, and synaptic plasticity in a cell-autonomous manner and thereby regulate sleep. Unfortunately, most experiments in the current study rely on a single genetic model (i.e., Mettl5 heterozygous mutant).

We believe that the multiple rescue experiments presented in Figure 1H-L and Figure 2H-L have effectively addressed the background concern. To further confirm this, we have subsequently repeated sleep and circadian rhythm assays using RNAi lines, aiming to further eliminate any remaining concerns in this regard. It appears to replicate the reduced sleep phenotype seen at night. This result has been included in the Figure S1. It is true that we have not specifically addressed whether the effects of Mettl5 are mapped to circadian pacemaker neurons in this study. We acknowledge this as a limitation and appreciate the importance of this question. Further investigations focusing on circadian pacemaker neurons, such as PDF- or tim-positive neurons, would be necessary to clarify the precise role of Mettl5 in regulating circadian behaviors and related molecular mechanisms.

(2) Gene expression and synaptic plasticity: gene expression profiles and the synaptic plasticity should be assessed by multiple time-point analyses since 1) they display high-amplitude oscillations over the 24-h window and 2) any phase-delaying mutation (e.g., Mettl5 deletion) could significantly affect their circadian changes. The current study performed a single time-point assessment of circadian clock/synaptic gene expression, misleading the conclusion for Mettl5 effects. Considering long-period rhythms in Mettl5 mutant clocks, transcriptome/translatome profiles in Mettl5 cannot distinguish between direct vs. indirect targets of Mettl5 (i.e., gene regulation by the loss of Mettl5-dependent m6A vs. by the delayed circadian phase in Mettl5 mutants).

In the revised version, we provided data collected at multiple time points. Specifically, we reexamined the per expression at both transcriptional and translational levels at different timepoints. The corresponding results were incorporated in Figure 4 D-F. We also dissected fly brains from UAS-DenMark, UAS-syt.eGFP/+; pdf-GAL4/+ and UAS-DenMark, UAS-syt.eGFP/+; pdf-GAL4/Mettl51bp at these four time points to quantify the synaptic structures of PDF neurons. The result has been included in revised Figure 6.

(3) The text description for gene expression profiling and Mettl5-dependent gene regulation was very detailed, yet there is a huge gap between gene expression profiling and sleep/behavioral analyses. The model in Figure 5 should be better addressed and validated.

Thank you for your suggestion. We added data to better confirm the expression changes of PER protein at different time points. Indeed, what you mention is the weak point of this paper. We did analysis thoroughly during the revision process.

The opposing changes in Period at the transcriptional versus translational levels puzzled us for some time until we identified alterations in the ubiquitin pathway components. The regulation of Period protein degradation by the ubiquitin-proteasome pathway is well-documented (Grima et al., 2002; Ko et al., 2002; Chiu et al., 2008). Additionally, previous studies have shown that N6-methyladenosine (m6A) modulates the ubiquitin-proteasome pathway in skeletal muscle physiology (Sun et al., 2023). We have incorporated this information into the revised manuscript in the last paragraph under the section titled: Clock gene regulatory loop regulating circadian rhythm was affected by Mettl51bp

Indeed, we have not yet identified an effective method to manipulate proteasome levels in genetic tests. The proteasome is a large protein complex composed of numerous subunits, making it impractical to enhance its activity simply by overexpressing individual components. Furthermore, the proteasome plays a critical role in many biological processes. Disrupting its function—such as through MG132 treatment, which we attempted—leads to significant off-target effects.

Sun J, Zhou H, Chen Z, et al. Altered m6A RNA methylation governs denervation-induced muscle atrophy by regulating ubiquitin proteasome pathway. J Transl Med. 2023;21(1):845. Published 2023 Nov 23. doi:10.1186/s12967-023-04694-3

Grima, B. et al. The F-box protein slimb controls the levels of clock proteins period and timeless. Nature 420, 178–182 (2002).

Ko, H. W., Jiang, J. & Edery, I. Role for Slimb in the degradation of Drosophila period protein phosphorylated by doubletime. Nature 420, 673–678 (2002).

Chiu, J. C., Vanselow, J. T., Kramer, A. & Edery, I. The phosphooccupancy of an atypical SLIMB-binding site on PERIOD that is phosphorylated by DOUBLETIME controls the pace of the clock. Genes Dev. 22, 1758–1772 (2008).

Reviewer #3 (Public review):

Xiaoyu Wu and colleagues examined the potential role in sleep of a Drosophila ribosomal RNA methyltransferase, mettl5. Based on sleep defects reported in CRISPR generated mutants, the authors performed both RNA-seq and Ribo-seq analyses of head tissue from mutants and compared to control animals collected at the same time point. While these data were subjected to a thorough analysis, it was difficult to understand the relative direction of differential expression between the two genotypes. In any case, a major conclusion was that the mutant showed altered expression of circadian clock genes, and that the altered expression of the period gene in particular accounted for the sleep defect reported in the mettl5 mutant. As noted above, a strength of this work is its relevance to a human developmental disorder as well as the transcriptomic and ribosomal profiling of the mutant. However, there are numerous weaknesses in the manuscript, most of which stem from misinterpretation of the findings, some methodological approaches, and also a lack of method detail provided. The authors seemed to have missed a major phenotype associated with the mettl5 mutant, which is that it caused a significant increase in period length, which was apparent even in a light: dark cycle. Thus the effect of the mutant on clock gene expression more likely contributed to this phenotype than any associated with changes in sleep behavior.

Recommendations for the authors:

Reviewer #1 (Recommendations for the authors):

Some of the questions that the authors should address are the following ones:

How does Mettl5 control the translation of the clock genes ? Why the level of some genes are specifically increased or decreased? What is the relation with the effect on uORF and dORF, overlapping and non overlapping ones? The observation of these defects is interesting but how they occurs and how they impact clock signaling is missing.

Thank you for your suggestion. This is the weak point of this paper. We did analysis thoroughly during the revision process.

The opposing changes in Period at the transcriptional versus translational levels puzzled us for some time until we identified alterations in the ubiquitin pathway components. The regulation of Period protein degradation by the ubiquitin-proteasome pathway is well-documented (Grima et al., 2002; Ko et al., 2002; Chiu et al., 2008). Additionally, previous studies have shown that N6-methyladenosine (m6A) modulates the ubiquitin-proteasome pathway in skeletal muscle physiology (Sun et al., 2023). We have incorporated this information into the revised manuscript in the last paragraph under the section titled: Clock gene regulatory loop regulating circadian rhythm was affected by Mettl51bp.

Indeed, we have not yet identified an effective method to manipulate proteasome levels in genetic tests. The proteasome is a large protein complex composed of numerous subunits, making it impractical to enhance its activity simply by overexpressing individual components. Furthermore, the proteasome plays a critical role in many biological processes. Disrupting its function—such as through MG132 treatment, which we attempted—leads to significant off-target effects.

In this study, we also observed codon usage alterations caused by the mettl5 mutant. For details, please refer to our responses to 4th question of the weakness session above. Previous studies have reported mettl5's role in translational regulation in other systems (Rong et al., 2020; Peng et al., 2022). Based on these findings, we propose that both translational regulation and protein degradation may contribute to the upregulation of Period protein in the mettl5 mutant. This hypothesis has been included in the Discussion section of the revised manuscript.

“The mechanism by which METTL5 regulates translation warrants further investigation. Previous studies have demonstrated that METTL5 influences translation (Rong et al., 2020; Peng et al., 2022), but whether the mechanisms identified here are conserved across other systems remains an intriguing question. In our analysis, we observed increased usage of aspartate (Asp) codons in Mettl5 mutants. Notably, prior work has linked codon usage to PER protein function—specifically, a codon-optimized version of PER failed to rescue circadian rhythmicity in per mutant flies, unlike the wild-type version (Fu et al., 2016). Further analysis revealed that PER protein levels were elevated in these mutants, suggesting that codon optimization enhances PER expression (Figure 2B in Fu et al., 2016). Strikingly, when we examined the codon-optimized region from Fu et al. (2016), we found that GAC (Asp) was highly enriched, raising the possibility that Mettl5 mutation affects PER protein accumulation by altering GAC codon usage. Additional experiments will be needed to validate this hypothesis. Furthermore, we detected changes in upstream open reading frames (uORFs) in Mettl5 mutants, but their relationship to translational regulation requires further exploration.”

References

Sun J, Zhou H, Chen Z, et al. Altered m6A RNA methylation governs denervation-induced muscle atrophy by regulating ubiquitin proteasome pathway. J Transl Med. 2023;21(1):845. Published 2023 Nov 23. doi:10.1186/s12967-023-04694-3

Grima, B. et al. The F-box protein slimb controls the levels of clock proteins period and timeless. Nature 420, 178–182 (2002).

Ko, H. W., Jiang, J. & Edery, I. Role for Slimb in the degradation of Drosophila period protein phosphorylated by doubletime. Nature 420, 673–678 (2002).

Chiu, J. C., Vanselow, J. T., Kramer, A. & Edery, I. The phosphooccupancy of an atypical SLIMB-binding site on PERIOD that is phosphorylated by DOUBLETIME controls the pace of the clock. Genes Dev. 22, 1758–1772 (2008).

Rong B, Zhang Q, Wan J, et al. Ribosome 18S m6A Methyltransferase METTL5 Promotes Translation Initiation and Breast Cancer Cell Growth. Cell Rep. 2020;33(12):108544. doi:10.1016/j.celrep.2020.108544

Peng H, Chen B, Wei W, et al. N6-methyladenosine (m6A) in 18S rRNA promotes fatty acid metabolism and oncogenic transformation. Nat Metab. 2022;4(8):1041-1054. doi:10.1038/s42255-022-00622-9

Fu J, Murphy KA, Zhou M, Li YH, Lam VH, Tabuloc CA, Chiu JC, Liu Y. Codon usage affects the structure and function of the Drosophila circadian clock protein PERIOD. Genes Dev. 2016 Aug 1;30(15):1761-75.

Reviewer #2 (Recommendations for the authors):

Please find my comments to improve the quality of your manuscript.

Major comments

(1) The quality of text writing in English needs to be at publishable levels. It is not a trivial problem, but it literally impairs the readability of your work. So please have professionals edit your manuscript text appropriately.

We have carefully revised the language throughout the manuscript during the revision process.

(2) Fig 1O: please include the total sleep profile and other analyses for rebound sleep phenotypes in control vs. Mettl5 to better validate that both genotypes were comparably sleep-deprived, but the latter shows less sleep rebound.

Thank you for your suggestion, The other reviewer also suggested to reanalyze the sleep rebound data. We did the analysis according to the following reference. We included data sleep profiles of both genotypes in original Fig 1O. Total sleep profile and other analyses for rebound sleep phenotypes are included in the revised panel. As shown in this revised panel (now Figure 1K, L), both genotypes were comparably sleep-deprived.

Cirelli C, Bushey D, Hill S, Huber R, Kreber R, Ganetzky B, Tononi G. 2005. Reduced sleep in Drosophila Shaker mutants. Nature 434:1087-92.

(3) Line 90: the authors did not actually address this critical question. Additional Gal4 mapping (e.g., Mettl5 rescue or Mettl5 RNAi) will determine which cells/neural circuits are important for Mettl5-dependent sleep.

This sentence has been revised into “The observed expression pattern of Mettl5 further supports its sleep regulatory function.”

(4) Fig 1H-L; Fig 2H-L: the authors should check if overexpression of wild-type or mutant Mettl5 in control backgrounds could affect nighttime sleep to better define the transgenic effects among overexpression, rescue, and dominant-negative.

Thank you for the comment. We added the overexpression phenotypes in the revised version.

(5) Lines 225-226. Fig S11: The neural projections from PDF-expressing neurons should be better imaged and quantified. Current images can visualize PDF projections onto the optic lobe but not others (e.g., dorsal, POT), so the conclusion is not validated.

Thank you for the suggestion. We acknowledge the limitation in the current images of PDF-expressing neuronal projections. We included new, higher-resolution images to better visualize and quantify the neural projections, including the dorsal and POT regions, to ensure the conclusion is well-supported.

(6) Lines 230-232: per RNA/PER protein expression oscillates daily, so the authors should perform time-point experiments to conclude Mettl5 effects on clock gene expression, including per.

Thank you for the insightful comment. We performed experiments in the Mettl5 mutant background at four time points to analyze PER protein expression using both RT-PCR and Western blot (anti-PER). The updated results have been included in Figure 4D-F.

(7) Lines 235-238: the authors should note that Mettl5 effects on sleep in Clk or per mutant backgrounds are actually opposite to those in w1118/control one. Mettl5 deletion promotes daytime or nighttime sleep in Clk or per mutants, respectively. Any explanation?

We are trying to use epistasis analysis to determine which gene is upstream here. Epistasis (or epistatic effect) in genetics refers to the interaction between different genes where the expression of one gene (the epistatic gene) masks or modifies the expression of another gene (the hypostatic gene). The epistatic gene (masking gene) usually functions downstream in the pathway because its effect overrides the output of the hypostatic gene. The double mutant showed the similar phenotype as downstream genes. Thus, Clk or per functions downstream of Mettl5.

(8) Fig 6: The dorsal PDF projections actually show time-dependent plasticity. Results from the single time-point are not conclusive.

Thank you for the insightful comment. we further dissected fly brains from UAS-DenMark, UAS-syt.eGFP/+; pdf-GAL4/+ and UAS-DenMark, UAS-syt.eGFP/+; pdf-GAL4/Mettl51bp at these four time points to analyze the morphology of PDF neurons. The results have been included in figure 6.

Minor comments

(1) Please avoid simple bar graphs in the data presentation-include individual data points or use a different graph showing the distribution of raw data (e.g., violin plot, box plot, etc.).

Thank you for the suggestion. In the revised version of the manuscript, we have included individual data points, violin plots, and box plots to present the data, effectively showing both the distribution and differences in the raw data.

(2) Line 19: "Clock" indicates the gene name or general terminology such as "circadian clock". Please clarify it and revise the font accordingly.

This has been revised into“clock”

(3) The overall flow in the Abstract/Summary is somewhat challenging for a general audience to follow.

We have revised the text, especially the overall flow in the Abstract/Summary.

(4) Fonts for the names of genes and gene products (i.e., mRNA, protein) should be appropriately corrected throughout the manuscript.

We have checked the text and made changes where necessary.

(5) Methods: the authors should provide detailed information on the methods. For instance, there is little description of how they generate Mettl5 deletions (e.g., sgRNA/target sequence). Also, they should clarify whether they test heterozygous vs. homozygous mutants of Mettl5 deletions in each experiment since the genotype description in the figure appears mixed-up (e.g., Fig 1B vs. Fig 1I-L).

Thank you for pointing this out. In the updated version, we provided detailed information about the strains used, including the sgRNA/target sequences for generating Mettl5 deletions. Regarding the genotypes, Figure 1B represents homozygous mutants, while Figures 1I-L represent heterozygous mutants. This distinction has been clarified in the figure legends, and the genotype notation for Figures 1I-L will be revised for consistency and clarity.

(6) Fig 1: the figure panels should be re-arranged based on the order of their text description (i.e., Fig 1H-L should go after Fig 1M-O).

Thank you for the suggestion. In the revised version, we rearranged the figure panels so that Figures 1H-L appear after Figures 1M-O, following the order of their description in the text.

(7) Sleep education in Trmt112 RNAi looks different from that in Mettl5 mutant het. Any explanation?

The functional divergence between Trmt112 and Mettl5 may also contribute to the observed sleep phenotype. While Trmt112 and Mettl5 share some downstream targets, they each regulate many unique genes, some of which could influence sleep. Sleep is a highly sensitive trait that can be modulated by numerous genetic factors. Previous studies have also suggested that sleep behaves more like a quantitative trait, reflecting the combined effects of multiple genes (Mackay and Huang, 2018).

Mackay TFC, Huang W. Charting the genotype-phenotype map: lessons from the Drosophila melanogaster Genetic Reference Panel. Wiley Interdiscip Rev Dev Biol. 2018;7(1):10.1002/wdev.289. doi:10.1002/wdev.289

Reviewer #3 (Recommendations for the authors):

A detailed critique is provided below. Generally, the authors can greatly improve this manuscript if they focus more rigorously on the circadian phenotype associated with the Mettl5 mutant, which could be the basis for the apparent sleep phenotype.

(1) Please provide more information as to how each of the mettl5 mutants were generated. This information should include, specifically, the gRNA sequences, plasmids generated for the 5' and 3' arms, and anything related to the CRISPR approach for generating the mutants. Was any sequencing done to verify the CRISPR alleles, or was this limited to the analysis of mettl5 expression and behavior? Please indicate where the qPCR primers (used in Fig 1B) are located relative to the mutant loci. The figure legend is also incomplete in that there is no reference to the boxed area in Fig 1A.

In the updated version, we have provided detailed information about the how each of the mettl5 mutants were generated. The sequence was verified by sequencing following PCR. The following references to the boxed area were added in the revised version.

Reference

Iyer LM, Zhang D, Aravind L. Adenine methylation in eukaryotes: Apprehending the complex evolutionary history and functional potential of an epigenetic modification. Bioessays. 2016 Jan;38(1):27-40. doi: 10.1002/bies.201500104.

(2) As noted, I am not in agreement with the interpretation of findings for the sleep defect reported in the mettl5[1b]/+ mutants. There is a clear increase in morning sleep in the mutants that may not have reached significance by lumping the data in 12h increments (Fig1C-E). Were the overall 24h sleep values between the mutants and controls the same? The sleep profile appears to be shifted, such that nighttime sleep onset in the mutants occurs much later than wild type, and daytime waking is also much later, all pointing to a long period phenotype, which is very strongly supported by the data in Table 1, as well as the RNA- and ribo-seq data. The implications for this leading to sleep disturbances in humans is very exciting. An additional suggestion to the authors here is to report the nighttime sleep latency values (time to onset of the first sleep bout after lights off).

We appreciate your insightful observation. As shown in Table 1, the Mettl51bp/+ mutant exhibits a robust long-period phenotype, with circadian rhythms significantly extended to 28.3 ± 0.4 hours compared to the wild-type's 23.9 ± 0.05 hours. This prolonged period perfectly aligns with the observed behavioral phenotypes, including delayed nighttime sleep onset, later daytime waking, and the overall shift in sleep profile. This is indeed quite similar to previous report on Period3 variant (Zhang et al., 2016). We agree that the prolonged circadian period contributes to the observed sleep phenotype. However, since total sleep time was significantly reduced in the mutant, we cannot attribute the phenotype solely to period lengthening. Furthermore, our 24-hour PER expression analysis in mettl5 mutants revealed elevated PER protein levels at ZT1 and ZT18, while ZT6 and ZT12 showed no significant changes, with no apparent phase shift. These findings collectively suggest that the phenotype primarily results from PER protein stabilization and accumulation.

Importantly, genetic rescue experiments restoring wild-type Mettl5 function (UAS-Mettl5/Mettl5-Gal4; Figure 1 and Table 1) completely normalized the circadian period to 24 ± 0.02 hours, providing compelling evidence that these phenotypes specifically result from loss of Mettl5 function. Together with the sleep architecture data, these findings establish Mettl5 as a crucial regulator of circadian rhythms, with important implications for understanding human sleep disorders. To further substantiate these observations, we have now included quantitative nighttime sleep latency measurements in the revised manuscript to better document the delayed sleep onset in mutants (Figure S1G).

We have discussed this in the third paragraph of the Discussion session and included the reference in the revised manuscript.

Zhang L, Hirano A, Hsu PK, et al. A PERIOD3 variant causes a circadian phenotype and is associated with a seasonal mood trait. Proc Natl Acad Sci U S A. 2016;113(11):E1536-E1544. doi:10.1073/pnas.1600039113.

(3) The description for how circadian behavior was measured and analyzed (Table 1) is missing from the methods section.

We have included a detailed description of the methods used to measure and analyze circadian behavior, as presented in Table 1, in the revised methods “Sleep behavior assays” section.

(4) Please explain what the "awake %" values reported in Figs 1G, 1L, Fig 2G, and 2L, Fig 4G and 4M are. Is this simply the number of flies that are awake at a given time point? This does not provide useful information beyond what is already reported for the sleep profiling in other parts of these figures. If it is an arousal threshold assay, as shown in supplementary Fig 1H, please indicate this. The description for "sleep arousal" in the methods (lines 368-371) is also concerning. If most of the mutant flies are already awake at ZT 14, then I would expect that this assay would not work at this time of day. A more suitable time point would be ZT 19, or later, when the mutants are falling asleep. Moreover, calculating the number of flies awakened as long as 5 minutes after a stimulus pulse cannot be distinguished from a spontaneous awakening, and so is not really a metric of arousal threshold. The number of sleeping flies awakened by the stimulus should be calculated within, at most, one minute afterward.

Thank you for your suggestion. Regarding the 'awake %' metric, it indicates that at specific time points (e.g., ZT14), the percentage of awake fruit fly population at that moment. In the revised version, we further clarify the definition and significance of 'awake %'. Additionally, we have reevaluated the time points for the arousal threshold assay, selecting a more appropriate time (e.g., ZT19) to better reflect the sleep state of the mutants. Based on your suggestion, we calculate the number of flies awakened within one minute after the stimulus to ensure a more accurate measurement of arousal threshold. This has been included in the revised Figure 1M.

(5) Fig1M-O is problematic. First, is it possible that expression of Mettl5 mRNA fluctuates with time-of-day and is not affected by sleep loss? There are no undisturbed controls collected at equivalent time points. The method used for quantifying sleep rebound in Fig 1O (lines 365-367) does not make sense, as negative values would be expected. Moreover, since the Mettl5 mutants show high sleep amounts in the morning and very low sleep amounts from ZT 12-18, this analysis would be severely confounded. Also, the sleep deprivation applied would not produce equivalent amounts of sleep loss as compared to wild type controls, so this also needs to be corrected. The authors should consider consulting Cirelli et al (2005, DOI: 10.1038/nature03486 ) as an approach for quantifying sleep homeostasis in a short-sleeping mutant. Please also show the sleep profiling in the mutants for these experiments.

Thank you for your valuable suggestions. Regarding the possibility that Mettl5 mRNA expression fluctuates with circadian rhythms rather than being affected by sleep deprivation, we acknowledge that collecting undisturbed control samples at equivalent time points would provide critical insights. In the revised version, we included undisturbed controls to distinguish between circadian-driven fluctuations and the effects of sleep deprivation on Mettl5 expression.

For the quantification of sleep rebound in Figure 1O, we agree that the current method may not fully capture the dynamics of sleep recovery, especially in Mettl5 mutants, where sleep patterns differ significantly from wild-type. We have referred to the method proposed by Cirelli et al. paper for quantifying sleep homeostasis in short-sleeping mutants, ensuring a more accurate evaluation of sleep rebound. The results have been included in Figure 1K-L of the revised version.

(6) Fig 3B and C (minor) - while the volcano plots are clear, it is not clear whether "down" or "up" means for the mutant relative to wild type or the other way around? Please clarify. In Fig 3P, the legend indicates a depiction of the "top 5 pathway associated genes", but it seems there are 10 pathways depicted. Which of these are the "top 5"?

In the volcano plots (Fig. 3B and 3C), “up” and “down” refer to genes that the mutant relative to the wild-type strain. In Fig. 3P, the legend was mislabeled as “top 5” pathway-associated genes. In fact, we displayed the top 10 pathway-associated genes. We apologize for the confusion and will correct both the figure legend and the corresponding text in our revised manuscript.

(7) Fig 4 D-E, and F,G do not have sufficient information to draw the conclusion that Per mRNA/protein expression is increased in the Mettl5 mutant. Since both mRNA protein of this gene oscillates significantly throughout the day, it is still possible that the single time point shown in this figure might indicate a disruption in cycling rather than overall expression level. Please first indicate what time of day the tissue was collected, second, consider adding more time points to both assays. For the first part of this figure, A and B, per and Clock gene expression are expected to be in different phases, and so this aspect is not unexpected. However, it is notable that it is reversed in the mutant vs wild type. Again, an alternate interpretation of this finding that the authors have not considered is a change in period duration of gene cycling.

Thank you for your suggestion. For the PER WB experiments, we have included multiple time points in the revised version to more comprehensively evaluate PER expression in the Mettl5 mutant and better understand its circadian rhythm changes. We appreciate your observation regarding the potential changes in the period duration of gene cycling. This has been discussed in the 3rd paragraph of the Discussion session of the revised version.

(8) The data shown in Figs 4H-M does not support the conclusion that "Clock and Per genes were downstream of Mettl5" (line 236-237). The daytime sleep phenotype, in particular, appears additive between both circadian genes and mutant because the morning sleep of the double mutant is much higher than either mutant by itself. Statistical comparisons between the double mutant and each clock mutant are also noticeably missing. These data are difficult to interpret. One potential explanation is that Mettl5 alters gene expression of non-circadian genes, and that the phenotypes become additive when both clock and Mettl5 genes are missing. A full molecular analysis of clock gene cycling in the Mettl5 mutant may help improve understanding of the relationship between the circadian clock Mettl5 gene expression. It may also be worthwhile checking whether Mettl5 gene expression itself shows a daily oscillation.

Thank you for your suggestion. In the revised version, we have included four additional time points to analyze the oscillatory expression of Per and Clock in the Mettl5 mutant, providing a more comprehensive understanding of their circadian rhythm changes. In Figs 4H-M, we are trying to use epistasis analysis to determine which gene is upstream here. Epistasis (or epistatic effect) in genetics refers to the interaction between different genes where the expression of one gene (the epistatic gene) masks or modifies the expression of another gene (the hypostatic gene). The epistatic gene (masking gene) usually functions downstream in the pathway because its effect overrides the output of the hypostatic gene. The double mutant showed the similar phenotype as downstream genes. Thus, Clk or per functions downstream of Mettl5. Statistical comparisons between the double mutant and each clock mutant are added.

(9) In Fig 6, what time of day were the flies collected? PDF terminal morphology is known to change throughout the day; this is another piece of data that could indicate a defect in circadian function rather than a chronic change in synaptic morphology.

The flies were collected around ZT14. We included additional dissection time points in future experiments. Differences between the control and Mettl5 mutants are observed consistently across multiple time points, suggesting that Mettl5 has an impact on synaptic plasticity.

Minor:

There are letter indicators, presumably for statistical comparisons, depicted in Figs 1 and 2 (panels I-L), but no explanation as to what these mean in the figure legends.

We have added notes in the revised version.

What is the purpose of the boxed regions shown in Fig S1A-F? There is no explanation of these in the figure legend nor in the text.

The boxed regions highlight the significant co-localization of two proteins. We have included this explanation in the figure legend in the revised version.

The statement (lines 310-311) that per and clock genes "exhibit more pronounced sleep rebound after sleep deprivation" is inaccurate. The article cited for this (Shaw et al 2002) showed that it was female mutants of the cycle gene which showed prolonged sleep rebound; other clock mutants were normal.

Thank you for pointing out this. We revised the statement accordingly.

Overall, the manuscript may benefit from editing or writing assistance to improve the language. There were many incomplete sentences, grammatical errors, etc.

We have carefully refined the language throughout the manuscript during the revision process.

Version 2

Abstract

Sleep plays a critical role in animal physiology, primarily governed by the brain, and its disruption is prevalent in various brain disorders. Mettl5 is associated with intellectual disability (ID), which often includes sleep disturbances. However, the mechanism underlying these sleep disruptions in ID remains poorly understood. In this study, we investigated the sleep phenotypes resulting from Drosophila Mettl5 mutations. Rescue experiments revealed that Mettl5 functions predominantly within neurons and glia marked by Mettl5-Gal4 to regulate sleep. Previous work established that Mettl5 forms a complex with Trmt112 to influence rRNA methylation. Notably, a mutation in Trmt112 mutation recapitulated these sleep disturbances, implicating translational regulation by the Mettl5/Trmt112 complex. Subsequent RNA-seq and Ribo-seq analyses of Mettl51bp mutants uncovered downstream effects, including altered expression of proteasome components and clock genes. Rescue experiments confirmed that the net increase in PERIOD protein underlies the sleep phenotype. This study illuminates the interplay between ribosome function, clock genes, and the proteasome in sleep regulation, highlighting the integrated roles of protein synthesis and degradation. These findings could potentially provide an example for in vivo study of rRNA methylation function, expand our understanding of protein homeostasis in sleep and offer insights into the sleep phenotypes associated with ID.

Introduction

Sleep is essential for animal physiology, and understanding its molecular mechanisms has significant implications for both basic research and clinical applications. Although previous studies have identified key sleep regulators (Du et al., 2021), fundamental questions remain regarding the roles of cellular processes such as protein synthesis and degradation in sleep regulation.

Growing evidence suggests a dynamic interplay between sleep regulation and protein homeostasis. Protein synthesis is particularly active during sleep (Lyons et al., 2023), while Sleep deprivation affects translational initiation (Costa et al., 2019). The proteasome, a critical mediator of ubiquitin-dependent protein degradation, regulates synapse homeostasis by modulating ribosomal components (Costa et al., 2019), thereby influencing synaptic function. Intriguingly, proteasome components exhibit oscillating expression pattern at the transcriptional level in human cells (Desvergne et al., 2016). A recent study in Drosophila showed that mutation in a proteasome component alters sleep patterns (Fernández et al., 2020). Despite these advances, the precise mechanisms by which protein synthesis and degradation influence sleep, and how these processes interact, remain unresolved.

While several factors influencing proteasome assembly and subunit transcription have been identified (Kapetanou et al., 2022), the translational control of proteasome subunits remains poorly understood. Additionally, proteasome inhibition has been shown to alter ribosome function (Galimberti et al., 2016; Costa et sl., 2019; Palanca et al., 2014), raising the question of whether the proteasome itself is regulated by the status of the translation machinery. Exploring this relationship could provide key insights into the interplay between proteasome activity and translation.

In Drosophila, Mettl5 exhibits 18S ribosomal RNA m6A methyltransferase activity (Leismann et al., 2020) and interacts with Trmt112 to facilitate this function, a mechanism conserved in human (Leismann et al., 2020; van Tran et al., 2019). Although Mettl5 depletion abolishes m6A modification on 18S rRNA without impairing rRNA maturation, it affects fly orientation behavior. In mammalian cells, Mettl5-mediated 18S rRNA N6-methyladenosine (m6A) modification regulates stem cell fate determination and neural function (Wang et al., 2020). However, it remains unclear whether Mettl5 modulates global or transcript-specific translation profile in vivo, particularly in tissues rather than cell lines. Moreover, while Mettl5 is linked to intellectual disability (ID) with comorbid sleep disturbances, its mechanistic role in these disorders is unknown. Testing whether Mettl5 loss affects behaviors like sleep in Drosophila could clarify this connection.

We discovered that Mettl5 regulates sleep in Drosophila. To investigate the underlying mechanism, we performed RNA-seq and Ribo-seq on Mettl5 mutants, revealing dysregulation of multiple clock genes and proteasome components. This suggests Mettl5 coordinates protein production and degradation, which are crucial for protein homeostasis. Follow-up experiments confirmed that the protein level of Period was upregulated in Mettl5 mutants, contributing to the sleep phenotype. This study highlights that ribosome defects can perturb proteasome function, uncovering a mechanism that couples protein degradation with synthesis. Additionally, by mapping the genome-wide downstream gene profile of an rRNA methylation modifier, this study offers insights into the gene-specific roles of ribosome function.

Results

Mettl5 regulates sleep in Drosophila

Mettl5 contains a predicted N6 adenine specific nucleic acids methyltransferase domain. In order to study the function of this gene in sleep regulation, we generated CRISPR-Cas9 knockout mutants. We created two alleles. Mettl51bp, which deleted 1bp in the coding region results in a truncated version (Figure 1A) and down regulation of Mettl5 at mRNA level (Figure 1B). Another allele Mettl59bp, which results in 3 amino acids deleted version of Mettl5 (Figure 1A) does not cause significant change at mRNA level (Figure 1B). We found that heterozygous Mettl5¹ᵇᵖ mutants exhibited significantly reduced nighttime sleep (Figure 1C-G), particularly during early night (ZT12-ZT16; Figure 1C). Quantitative analysis revealed increased wakefulness at ZT14 in mutants (Figure 1H). ’awake %’ was used to indicate the percentage of awake fruit fly population at specific time points (e.g., ZT14). quantitative nighttime sleep latency measurements indicated a delayed sleep start in mutants (Figure S1G). In addition, Mettl5 expression was significantly up regulated during the sleep recovery period after mechanical sleep deprivation (Figure 1I-J). Mettl5 mutant displayed significantly increased sleep rebound in 24 hours after sleep deprivation (Figure 1K-L), indicating its effects on sleep homeostasis. Moreover, results of sleep arousal assay at ZT19 indicated that the percentage of aroused flies are significantly more than the control group (Figure 1M). These results demonstrate that the truncated Mettl5 mutation causes sleep deficits, establishing Mettl5 as a novel sleep regulator in Drosophila.

Mettl5 is a regulator of Drosophila sleep.

(A) Diagram illustrating CRISPR-Cas knockout of 1 or 9 bases in the Mettl5 gene. (B) Relative expression of Mettl5 mRNA in homozygous Mettl51bp and Mettl59bp mutant male flies compared to control flies. (C) Sleep curve throughout the day for Mettl5 mutant male flies (blue) and control flies (black). (D) Total sleep of Mettl5 mutant male flies and control flies in 24 hours. (E) Total sleep of Mettl5 mutant male flies and control flies within day and night, respectively. (F) Sleep bout duration of Mettl5 mutant male flies and control flies. (G) Number of sleep bouts of Mettl5 mutant male flies and control flies. (H) Percentage of awake for Mettl5 mutant flies and control flies. (I) Sleep curve is tracked throughout the entire day prior to sleep deprivation and during the daytime sleep rebound period. (J) Mettl5 mRNA expression level at different time points. W(wake), SD (sleep deprivation), SR (sleep recovery). (K) Sleep curve is tracked throughout the entire day prior to sleep deprivation and during the daytime sleep rebound period in Mettl5 mutant male flies (blue) and control flies (black). (L) Response to sleep deprivation and performance measures in Mettl5 mutants and controls. Black bars represent the amount of sleep lost during the 24-hour sleep deprivation period, blue bars indicate the amount of sleep regained, whereas the red bars indicate the proportion of sleep recovered (right y-axis). (M) Sleep arousal of Mettl51bp male flies and control flies at ZT19. (N) Sleep curve throughout the day for the following genotypes: w1118(black), Mettl51bp/+ (blue), and Mettl5-Gal4, UAS-Mettl5, Mettl51bp/+(pink). (O) Total sleep of the indicated genotypes. (P) Sleep bout duration of the indicated genotypes. (Q) Number of sleep bouts of the indicated genotypes. (R) Percentage of awake for the indicated genotypes. For * stands for p<0.05, ** stands for p<0.01, *** stands for p<0.001, ns stands for not significant. For letter-based annotations, Groups with no significant differences share the same letter, Groups with significant differences are assigned new letters.

To confirm the specificity of the sleep phenotype in Mettl51bp, we performed genetic rescue experiments. Introduction of a single wild-type Mettl5 copy completely rescued both the reduced sleep amount and increased wakefulness at ZT14 (Figure 1N-R), demonstrating that these phenotypes specifically result from Mettl5 deficiency.

The observed expression pattern of Mettl5 further supports its sleep regulatory function. Using Mettl5-Gal4 reporter lines, we found expression in both neurons (colocalizing with ELAV staining; Figure S1A-C) and glial cells (colocalizing with Repo staining; Figure S1D-F). Behavioral analyses revealed additional mutant phenotypes consistent with sleep dysregulation. we tested the climbing ability of Mettl5 mutation which showed an increase (Figure S1H). RNAi knocking down of Mettl5 showed a consistent phenotype of down regulated sleep amount during the night time (Figure S1I-L).

Mettl5 regulates Drosophila sleep through its methyltransferase activity

Previous studies established that Mettl5 interacts with Trmt112 to regulate 18S rRNA m6A modification (Leismann et al., 2020; van Tran et al., 2019). Consistent with these findings, our LC-MS/MS analysis revealed significantly reduced m⁶A levels in both total RNA and 18S rRNA from heterozygous Mettl51bp (Figure 2A-B). To determine whether Mettl5’s sleep regulatory function depends on its methyltransferase activity, we performed two key experiments. First, we examined the phenotype of Trmt112 knockdown. We found that the Mettl5-Gal4 driven Trmt112 RNAi resulted in a similar phenotype to Mettl51bp (Figure 2C-G), suggesting that Mettl5 regulates sleep through its 18S rRNA m6A modification activity. The efficiency of the Trmt112 RNAi line was tested in previous study (López-Varea et al., 2021). More important, we performed a rescue experiment with a mutated form of Mettl5 that lacks the NPPF amino acids required for its methyltransferase activity (Figure2A) (Iyer et al., 2016). We observed that this mutant Mettl5 failed to rescue the sleep phenotype caused by Mettl51bp (Figure 2H-L). Together, these results demonstrate that Mettl5’s methyltransferase activity is essential for its role in sleep regulation, likely through its function in 18S rRNA m⁶A modification.

Mettl5 regulation of Drosophila sleep was dependent on its methyltransferase activity.

(A) The m6A level in the total RNA of Mettl5 mutant male flies. (B) The m6A level in the 18S rRNA of Mettl5 mutant male flies. (C) The sleep curve throughout the day shows the sleep pattern of induced Trmt112 RNAi male flies and control flies. (D) Total sleep of induced Trmt112 RNAi male flies and control flies. (E) Sleep bout duration in induced Trmt112 RNAi male flies and control flies. (F) Number of sleep bouts in induced Trmt112 RNAi male flies and control flies. (G) Percentage of awake in Trmt112 RNAi and control flies. (H) Sleep curve throughout the day for Mettl5 mutant male flies, induced Mettl5m over expression male flies and control flies. (I) Total sleep of Mettl5 mutant male flies, induced Mettl5m over expression male flies and control flies. (J) Sleep bout duration in Mettl5 mutant male flies, induced Mettl5 over expression male flies and control flies. (K) Number of sleep bouts in Mettl5 mutant male flies, induced Mettl5 over expression male flies and control flies. (L) Percentage of awake in Mettl5 mutant male flies, induced Mettl5 over expression male flies and control flies. For * stands for p<0.05, ** stands for p<0.01, *** stands for p<0.001, ns stands for not significant. For letter-based annotations, Groups with no significant differences share the same letter, Groups with significant differences are assigned new letters.

RNA-seq and Ribo-seq revealed the downstream gene profile of Mettl5

To better understand the downstream events of Mettl5, we performed RNA-seq and Ribo-seq to assess transcriptomic and translational changes in Mettl51bp. Principal Coordinates Analysis (PCoA) of RNA-seq and Ribo-seq datasets revealed clear separation between mutant and control groups (Figure S2A, B, D, E). High reproducibility was found among biological replicates (Figure S2C, F), confirming data quality. Transcriptome analysis identified 1,053 significantly differentially expressed genes (|log2 (Fold Change) | ≥ 1 & padj< 0.05), comprising 217 upregulated and 836 downregulated transcripts (Figure 3A, B). Parallel ribosome profiling revealed 299 translationally regulated genes (|log2 (Fold Change) | ≥ 0.265 & padj< 0.05), with 149 upregulated and 150 downregulated targets compared to the controls (Figure 3C). Heatmaps of the top 100 differentially expressed genes from both RNA-seq and Ribo-seq analyses were generated according to the rlog transformed values (Figure S3A, B).

RNA-seq and Ribo-seq analysis revealed changes in the gene profile of Mettl51bp.

(A) Venn diagram depicting the number of significant differentially expressed genes revealed by RNA-seq and Ribo-seq. (B) Volcano plot representing the differentially expressed genes identified by RNA-seq. Genes that met the criteria of |log2(fold-change)| ≥ 1 and p.adjust < 0.05 were considered significantly expressed, marked in orange for downregulation and green for upregulation, comparing with the controls. (C) Volcano plot representing the differentially expressed genes identified by Ribo-seq. Candidates that satisfied the criteria of |log 2 (fold-change) | ≥ 0.265 and p.adjust <0.05 were regarded as significantly expressed, marked in red for downregulation and blue for upregulation, respectively. (D-K) Gene set enrichment analysis of differentially expressed genes revealed by RNA-seq (D-G) and Ribo-seq (H-K). All the plots are generated using the KEGG gene set database. The bar chart at the bottom of each panel shows the distribution of target genes for each pathway according to their rank position. Each vertical line represents a gene. Genes on the left show positive correlation with Mettl51bp, while, genes on the right show negative correlation with Mettl51bp. The green line indicates the enrichment score (ES), and NES stands for normalized enrichment score. (L) Distribution of the differentially expressed genes revealed by both RNA-seq and Ribo-seq. (M) Cumulative distribution of TE frequencies among w1118 and Mettl51bp (N, O) GO and KEGG enrichment of significantly changed TE-related genes between w1118and Mettl51bp. The color of the bar indicates the enrichment p.adjust value. (P) KEGG network showing the top 10 pathway and associated genes. The size of the dots represents the number of genes in the pathway.

To explore the biological implication of differentially expressed genes at transcriptional and translational levels, we performed Gene Ontology (GO) enrichment analysis. At the transcriptional level (Figure S4A, C), the most significantly enriched biological processes included cellular response to chemical stimulus, small molecule biosynthetic process, and fatty acid metabolic process (Figure S4A), with supporting gene networks (Figure S4C). Translational-level analysis (Figure S4B, D) revealed prominent enrichment for organic acid and amino acid metabolic processes, lipid catabolism, cellular respiration, and transport mechanisms. Notably, we identified strong associations with circadian regulation, including circadian regulation of gene expression and entrainment of the circadian clock.

The GO results usually contain a long list of enriched terms that have highly redundant information and are difficult to summarize. So, we performed a simplify Enrichment analysis that visualizes the summaries of clusters by word cloud for the GO enrichment result at transcriptional and translational levels. For transcriptionally significant DEGs (Figure S5A), major enriched processes included metabolic and stimulus-response pathways, wing disc and imaginal system development, ion-channel homeostasis and transport, and cytokinesis/cell cycle transitions. Translationally regulated genes additionally showed enrichment for sleep-wake cycles, circadian behavior, and neuronal cell death (Figure S5D). Cellular component analysis highlighted chromosomal structures, membrane-bound complexes, and vesicles (Figure S5B), while molecular functions predominantly involved enzymatic activities, ion/protein binding, receptor activities, and transcription processes (Figure S5C). The strong concordance between RNA-seq and Ribo-seq enrichment profiles (Figure S5E-F) further validated these findings.

Our Gene Set Enrichment Analysis (GSEA) identified pathways enriched in differentially expressed genes at both transcriptional and translational levels. To capture subtle but biologically important changes that might be excluded by stringent statistical thresholds, we analyzed gene sets ranked by log2 fold-change without p-value filtering. The GSEA-GO and GSEA-KEGG results, sorted by Normalized Enrichment Score (NES), revealed several key pathways (Figure S6A-D). Notably, the proteasome pathway showed significant suppression at both transcriptional and translational levels (P.adjust<0.05) (Figure 3E, I). Transcriptional-level analysis highlighted disruptions in arginine and proline metabolism and insect hormone biosynthesis (P.adjust<0.05) (Figure 3D, G). While, while translational changes predominantly affected fatty acid degradation and phosphonate/phosphinate metabolism (P.adjust<0.05) (Figure 3H-K).

To better understand the correlation between transcription and translation levels, we performed correlation analysis of differentially expressed genes based on fold change of DEGs (Figure 3L). 977 genes showed opposing trends between transcription and translation, while only 24 genes exhibited concordant changes. We identified 240 translation-specific and 3,309 transcription-specific DEGs, with each cluster undergoing separate enrichment analysis (Figure S7A-H). Notably, core clock genes (Clk, tim, etc.) displayed significant changes at both levels (Figure 3L).

To isolate translational effects, we calculated translation efficiency (TE) differences between Mettl51bp and w1118controls (Figure 3M). Among 1,204 genes with significantly altered TE, GO enrichment highlighted amino acid metabolism and small molecule biosynthesis (Figure 3N), while KEGG analysis emphasized glycine/serine/threonine metabolism and one-carbon pool by folate pathways (Figure 3O, P).

Ribo-seq revealed the Mettl51bp led to changes of some global translation features

Using Ribo-seq, we compared global translation features between the w1118 and Mettl51b groups. The TE (translation efficiency) correlation coefficients among the three biological replicates of each genotype ranged from 0.87 to 0.93, demonstrating high reproducibility within genotypes (Figure S8A-F). We found that the length of ribosome-protected RNA fragments (RPFs) was approximately 28nt (Figure S10A). The RPFs exhibited a significant 3-nt periodicity. Metagene analysis of individual 28nt reads revealed the distribution of RPFs across the gene locus in w1118 and Mettl51bp, respectively (Figure S8, S9). The starting point of translation is 12nt upstream of start-codon and gradually disappears 15nt from stop-codon (Figure S8G-L). RPFs on the metagene plot distribution around the translation start and translation stop site (Figure S9A-F), different coding frames on CDS, 3’-UTR and 5’-UTR, respectively in two groups of samples all showed the periodicity (Figure S9G-L). As expected, all plots show an enrichment of P-sites in the first frame on the coding sequence but not the UTRs, in accord with ribosome protected fragments from protein coding mRNAs (Figure S9G-L).

Analysis of open reading frame (ORF) types revealed alterations in their distribution. Compared to w1118, Mettl51bp exhibited an increased proportion of overlapping downstream ORFs (dORFs) and fewer non-overlapping dORFs (Figure S10B). In both groups, translated upstream ORFs (uORFs) were significantly shorter than untranslated uORFs (P < 0.05) (Figure S10C). Additionally, Mettl51bp displayed reduced read counts for both dORFs and uORFs (Figure S10D-E). Based on translation potential, uORFs were classified as either translated or untranslated, and motif analysis was performed separately for each category (Figure S10F-G).

Mutation of Mettl5 altered codon preference

Comparison of codon occupancy (A-site) between the two groups revealed that Mettl51bp preferentially used GAC and GAU, whereas w1118 favored UCC (Figure S10H). This trend was further supported by the cumulative frequency distribution of these codons (Figure S10I-K). Since GAC and GAU both encode aspartate (Asp), we analyzed Asp amino acid occupancy. Intriguingly, Asp was significantly enriched in w1118 during translation (Figure S10L), suggesting that Mettl51bp may exhibit altered translational regulation (see Discussion for details).

Metagene analysis of RPFs revealed distinct translation patterns between Mettl51bp and w1118. The coding sequence (CDS) and flanking regions were segmented into 100 equal bins, and average RPF density was computed for each bin. The resulting plots illustrate differences in ribosome occupancy between Mettl51bp and w1118 along the CDS (Figure S10M), near the translation start site (Figure S10N) and around the translation termination site (Figure S10O). Notably, Mettl51bp and w1118 exhibited divergent ribosome occupancy patterns, particularly along the CDS and near the start codon (Figure S10M, N), suggesting potential differences in translation dynamics and initiation efficiency.

Mettl5 regulates the clock gene regulatory loop

Our findings demonstrate that Mettl51bp disrupts the core clock gene regulatory loop controlling circadian rhythm. We observed significant alterations in both transcriptional and translational levels of multiple clock genes, with Cry and Clock showing upregulation while Tim, per, vri, and pdp1 were downregulated at both levels (Figure 4A-B). Notably, per, vri, and pdp1 exhibited particularly pronounced downregulation in translation efficiency (Figure 4C). These changes occurred without affecting clock neuron morphology at different time points (Figure S11).

Clock genes expression mediated the sleep phenotype caused by Mettl5 mutation.

(A-C) Fold changes in clock genes with significant expression level differences between w1118 and Mettl51bp were observed in RNA-seq, Ribo-seq, and translation efficiency analyses. (D) The gene expression levels of per at 4 different timepoints of w1118 and Mettl51bp. (E) Western blot analysis of PER protein and Tubulin protein at 4 different timepoints, (F) along with their ratio quantification. (G) Representative image of PER protein immunofluorescence staining in the small ventral lateral neurons (small LNvs). (H) Statistical analysis of the immunofluorescence intensity for PER in small LNvs. (I) Sleep curve throughout the day for Mettl51bp, per01, double mutant and control flies. (J) Total sleep for Mettl51bp, per01, double mutant and control flies. (K) Percentage of awake time in Mettl51bpflies, partially rescued by double mutant flies. (L) Sleep curve throughout the day for Mettl51bp, per01, double mutant and control flies. (M) Total sleep for Mettl51bp, per01, double mutant and control flies. (N) Percentage of awake time in Mettl51bp flies, partially rescued by double mutant flies. (O) Fold changes in proteasome subunits with significant expression level differences between w1118 and Mettl51bp were observed in RNA-seq, Ribo-seq, and translation efficiency analyses. For statistical significance, * stands for p<0.05, ** stands for p<0.01, *** stands for p<0.001, ns stands for not significant.

The observed expression patterns revealed an unexpected regulatory relationship: while the canonical clock circuitry positions Per downstream of Clock, our finding that Clock was upregulated while per was downregulated suggests Per may actually function upstream of Clock in Mettl5-mediated regulation. Surprisingly, despite the transcriptional downregulation of per (Figure 4D), we detected increased PER protein levels through both immunostaining and western blot analyses (Figure 4E-H). Detection of the PER protein at different time points indicated that it was increased at both ZT 0 and ZT18 (Figure 4E-F). This apparent contradiction aligns with the observed circadian phenotype, as Mettl5 mutants showed significantly longer period lengths (Table 1), mirroring effects seen when PER stabilization results from reduced kinase activity as previously reported (Philpott et al., 2023). Genetic epistasis experiments further supported this model, with double mutants exhibiting clock gene phenotypes that position both Clock and Per downstream of Mettl5 (Figure 4I-N, Table 1). Together, these results indicate that Mettl51bp affects circadian regulation through mechanisms that extend beyond transcriptional control, likely involving post-translational regulation of PER protein stability.

To investigate the factors mediating PER protein level changes in Mettl5 mutants, we examined the ubiquitin-proteasome pathway, which plays a well-documented role in Period protein degradation (Grima et al., 2002; Ko et al., 2002; Chiu et al., 2008). Notably, recent evidence indicates that N6-methyladenosine (m6A) regulates the ubiquitin-proteasome system in other biological contexts (Sun et al., 2023). Our integrated analysis of RNA-seq and Ribo-seq data revealed significant downregulation of multiple proteasome pathway components in Mettl51bp mutants at both transcriptional and translational levels (Figures 3E, 3I, 4O), suggesting impaired protein degradation capacity. Based on these findings, we propose a model where Mettl5 regulates circadian function through three interconnected mechanisms: first, by directly modulating proteasome components to control PER protein stability post-translationally; second, by transcriptionally and translationally regulating Per and other clock expression. In this model, Mettl51bp-induced proteasome downregulation leads to PER accumulation, which is responsible for the phenotypes (Figure 5).

A working model illustrating the role of Mettl5 in Drosophila sleep was presented.

Mettl51bp alters axon complexity

Previous studies have established a strong correlation between sleep homeostasis and synaptic complexity (Bushey et al., 2011). Additionally, synaptogenesis has been shown to enhance proteasome activity in axons (Costa et al., 2019). Given our observations of impaired sleep rebound following deprivation and altered expression of proteasome subunits in Mettl51bp mutants, we sought to examine potential effects on synaptic complexity.

To assess synaptic complexity, we adapted an established quantification method (Bushey et al., 2011) using Syt-GFP, a marker that colocalizes with endogenous synaptic vesicles, to visualize presynaptic morphology changes. Control experiments with UAS-Fmr1 and Fmr1 mutations successfully replicated the expected decrease and increase in syt-eGFP signal, respectively (Figure 6A-D). Strikingly, Mettl51bp mutants exhibited significantly increased syt-eGFP fluorescence in presynaptic terminals (Figure 6E-M), indicating altered synaptic complexity.

The axon complexity was found to be affected by Mettl51bp

(A–C) Representative confocal micrographs of small ventral lateral neuron (s-LNv) axonal terminals across different genotypes. Presynaptic structures were visualized using syt-eGFP (green), which colocalizes with endogenous synaptic vesicles: (A) control (B) Fmr1 overexpression and (C) Fmr1 null mutant. (D) Quantification of s-LNv axonal terminal volumes corresponding to genotypes in (A– C). (E–H) Representative images of axonal terminal morphology in control flies at four time points: ZT0, ZT6, ZT12, and ZT18. (I–L) Representative images of axonal terminal morphology in Mettl5 mutant flies at ZT0, ZT6, ZT12, and ZT18. (M) Quantitative comparison of axonal terminal volumes between control and Mettl5 mutant flies at different time points. Scale bar: 50 μm. For statistical significance, * stands for p<0.05, ** stands for p<0.01, *** stands for p<0.001, ns stands for not significant.

Discussion

Our study reveals that Mettl5, a known rRNA methyltransferase, modulates sleep through its RNA methylation activity. Through integrated RNA-seq and Ribo-seq analyses of Mettl51bpmutants, we identified Mettl5’s downstream targets at both transcriptional and translational levels. Further investigation demonstrated that Mettl5 influences sleep regulation by affecting two key pathways: the circadian clock gene network and the proteasome system. These findings provide novel mechanistic insights into sleep control, highlighting the coordinated role of protein synthesis and degradation in this process. Notably, our Ribo-seq analysis revealed that Mettl51bp alters fundamental translation features, including uORF translation efficiency and codon preference, suggesting rRNA methylation plays a regulatory role in these processes.

This discovery has important clinical implications, as METTL5, the human ortholog of Mettl5, is associated with intellectual disability (ID) when mutated (Richard et al., 2019). Our work expands the understanding of Mettl5’s molecular function and may inform potential therapeutic strategies for ID. The clinical relevance of our findings is underscored by reports of sleep disturbances, particularly reduced sleep duration, in ID patients - a phenotype that parallels our observations in Mettl51bp mutants. The mechanistic framework established in this study could explain these clinical sleep abnormalities. However, further validation in vertebrate models is needed to determine whether this regulatory mechanism is evolutionarily conserved and applicable to human sleep disorders.

As shown in Table 1, the Mettl51bp/+ mutant exhibits a robust long-period phenotype, with circadian rhythms significantly extended to 28.3 ± 0.4 hours compared to the wild-type’s 23.9 ± 0.05 hours. This prolonged period perfectly aligns with the observed behavioral phenotypes, including delayed nighttime sleep onset, later daytime waking, and the overall shift in sleep profile. This is indeed quite similar to previous report on Period3 variant (Zhang et al., 2016). We think that the prolonged circadian period contributes to the observed sleep phenotype. However, since total sleep time was significantly reduced in the mutant, we cannot attribute the phenotype solely to period lengthening. Furthermore, our 24-hour PER expression analysis in Mettl5 mutants revealed elevated PER protein levels at ZT1 and ZT18, while ZT6 and ZT12 showed no significant changes, with no apparent phase shift. These findings collectively suggest that the phenotype primarily results from PER protein stabilization and accumulation.

We found that Mettl5 heterozygotes showed significant reductions in total RNA and 18S rRNA methylation levels, contrasting with mouse studies where heterozygous knockouts maintained normal 18S rRNA m6A methylation (Sepich-Poore et al., 2022). This discrepancy may stem from either fundamental differences in rRNA methylation regulation between Drosophila and mice, or distinct biological consequences of knockdown versus knockout approaches, as complete gene elimination often triggers compensatory mechanisms (Teng et al., 2013; Rossi et al., 2015; Vu et al., 2015; Ma et al., 2019; El-Brolosy et al., 2019).

Our study uncovers a previously unrecognized connection between circadian clock genes and proteasome function. While previous work demonstrated that the circadian clock rhythmically regulates proteasome components in Drosophila fat bodies under dietary restriction (Hwangbo et al., 2023), we now show that Mettl5 modulates clock protein synthesis and degradation in clock neurons by influencing proteasome activity. This regulation likely occurs through Mettl5-mediated ribosomal methylation in clock neurons, which impacts the proteasome degradation pathway (Costa et al., 2019), ultimately altering clock protein dynamics.

Mettl5 represents a novel integrator of rRNA methylation and proteasome function, providing a mechanism to balance protein synthesis and degradation. Interestingly, we observed specific effects on both transcriptional and translational outputs, with particular proteasome subunits showing differential regulation. This specificity may arise from either selective translational control by Mettl5 or additional layers of regulation through protein-protein interactions. These findings suggest an intricate regulatory network coordinating these processes, though further studies are needed to elucidate the underlying mechanisms.

Our results reveal complex relationships between clock genes and sleep regulation. While cyc loss-of-function alleles show enhanced sleep rebound in females (Shaw et al., 2002), Mettl51bpmutants with elevated PER protein exhibit reduced rebound. Notably, Mettl5’s regulatory effects display tissue specificity - although absent in canonical clock neurons, Mettl5-Gal4 is expressed in distinct neurons and glia that appear crucial for sleep regulation. This expression pattern, combined with Clock’s broader distribution (Patop et al., 2023), suggests specialized circuits for sleep homeostasis that warrant further investigation.

Evidence indicates that sleep functions in development, metabolism and neuronal plasticity (Anafi et al., 2019). Sleep exerts effects on neuronal plasticity by modifying synapses. The synaptic homeostasis theory proposes that sleep has a role in downscale synaptic strength (Tononi and Cirelli, 2006). Indeed, synapse markers progressively decrease during sleep in both mammals and Drosophila brains (López-Varea et al., 2021; Vyazovskiy et al., 2008; Gilestro et al., 2009; Liu et al., 2010) More importantly, evidence indicates that sleep need and protein levels in synapse are tightly linked. Studies of the presynaptic active zone have shown that synaptic plasticity regulates sleep homeostasis (Huang et al., 2020). Genome-wide proteomic studies of synapse in mouse brain indicate that the synaptic proteins peak around dusk and dawn (Noya et al., 2019; Brüning et al., 2019). Sleep deprivation experiments have demonstrated that the sleep drive is significantly dependent on the cycling of proteins and phosphoproteins in synapses, in contrast to mRNAs (Noya et al., 2019; Brüning et al., 2019). These results suggest that the gene expression regulation at the protein level is crucial for sleep. Consistently, we detected changes in the quantity of axons in Mettl51bp, suggesting a possible regulation of neural circuits by Mettl5.

The mechanism by which METTL5 regulates translation warrants further investigation. Previous studies have demonstrated that METTL5 influences translation (Rong et al., 2020; Peng et al., 2022), but whether the mechanisms identified here are conserved across other systems remains an intriguing question. In our analysis, we observed increased usage of aspartate (Asp) codons in Mettl5 mutants. Notably, prior work has linked codon usage to PER protein function—specifically, a codon-optimized version of PER failed to rescue circadian rhythmicity in per mutant flies, unlike the wild-type version (Fu et al., 2016). Further analysis revealed that PER protein levels were elevated in these mutants, suggesting that codon optimization enhances PER expression (Figure 2B in Fu et al., 2016). Strikingly, when we examined the codon-optimized region from Fu et al. (2016), we found that GAC (Asp) was highly enriched, raising the possibility that Mettl5 mutation affects PER protein accumulation by altering GAC codon usage. Additional experiments will be needed to validate this hypothesis. Furthermore, we detected changes in upstream open reading frames (uORFs) in Mettl5 mutants, but their relationship to translational regulation requires further exploration.

Our study demonstrates that ribosomal components can exert gene-specific regulatory functions, building upon previous work showing context-dependent ribosome specialization (Simsek et al., 2017). In Mettl5 mutants, we observed distinct alterations in the translational efficiency of specific genes. These effects could potentially arise through two non-exclusive mechanisms: (1) Mettl5-mediated rRNA modifications may modulate ribosomal binding affinity for particular mRNA sequences, or (2) these modifications might contribute to the formation of specialized ribosome populations that preferentially translate specific subsets of mRNAs. Future studies will be required to distinguish between these possibilities and fully elucidate the underlying molecular mechanisms.

Material and methods

Drosophila Strains

Fly stocks used in this study were maintained under standard culture conditions. We used the w1118 as the control strain. The following flies were obtained from Bloomington Stock Center: w1118 (Bl: 5905), Mettl5-Gal4 (Bl: 19514), Repo-Gal4 (Bl: 7415), nSyb-Gal4 (Bl: 51941), pdf-Gal4 (Bl: 41286), ClkJRK (Bl: 80927), per01(Bl: 80917), Fmr1Δ50M(Bl: 6930), UAS-Trmt112RNAi (VDRC: 101515), UAS-DenMark, UAS-syt.eGFP (Bl: 33064), UAS-GFPstinger (from Yi Rao’s lab), UAS-Fmr1 (Bl: 6931), UAS-Mettl5 (FlyORF: F000760) are from Fly ORF collection. UAS-Mettl5m-3HA (generated in this study).

Mettl51bp and Mettl59bp mutants were generated by the CRISPR/Cas system as described previously (Cheng et al., 2020). The target gRNA was designed with an online tool: http://tools.flycrispr.molbio.wisc.edu/targetFinder/. A target sequence was chosen that has the sequence of 5’-GAGTGGGGTACTGTTCCAACAGG with the PAM sequence in bold. Mutations were verified by genomic PCR and sequencing.

Sleep behavior assays

All sleep assays were conducted in a controlled environment incubator maintained at 25 ± 1°C with 60% ± 5% relative humidity. We maintained a 12:12 light-dark cycle with lights on at ZT0 (06:30) and off at ZT12 (18:30). Fly activity was monitored using the Drosophila Activity Monitoring System (Trikinetics, Waltham, MA). Following a 2-day acclimation period, we recorded locomotor activity at 1-minute intervals for 3 consecutive days. Data analysis was performed using pysolo (Gilestro and Cirelli, 2009), with sleep defined as ≥5 minutes of continuous inactivity.

Mechanical sleep deprivation was performed using the SNAP method to keep flies awake for 12 h overnight (Shaw et al., 2002). Sleep deprivation is applied by mechanical stimuli using a timer-controlled rotating shaker. The intensity of stimuli was 1500 rpm once every minute (2 s/1 min) (Shimizu et al., 2008). Baseline sleep was established from 24-hour recordings prior to deprivation. Sleep loss and recovery were quantified according to previous publication (Cirelli C et al., 2005).

Regarding the ’awake %’ metric, it indicates that at specific time points (e.g., ZT14), the percentage of awake fruit fly population at that moment. At ZT19, we evaluated arousal thresholds by administering a standardized gentle mechanical stimulus. Responsiveness was determined by monitoring activity for 1 minutes post-stimulation. Flies showing no activity during this window were scored as non-responsive. We calculated arousal percentages from the proportion of flies that awakened in response to stimulation.

Circadian rhythm of individual male flies was measured using the Drosophila Activity Monitoring (DAM) System (Trikinetics). Male flies were loaded individually into glass tubes with a length of 65 mm and an inner diameter of 5 mm. The tubes contained standard cornmeal fly food at one end and were sealed with a cotton stopper at the other end. The flies were entrained to a 12 h light/12 h dark cycle for 3 days and then released to constant darkness for at least six days to measure their rhythmicity.Data analysis is done on a Macintosh computer running the FaasX (Fly activity analysis suite) software.

Statistical analysis

All statistical analyses were performed using GraphPad Prism 5 software. For sleep parameter comparisons, we used nonparametric tests including the two-tailed Mann-Whitney test for pairwise comparisons and one-way ANOVA with Tukey’s post hoc test for multiple comparisons. qPCR data were analyzed using unpaired Student’s t-tests, while lifespan data were evaluated with Log-rank (Mantel-Cox) tests. The specific statistical test used for each experiment is indicated in the corresponding figure. In all analyses, a p-value of less than 0.05 was considered statistically significant.

Quantitative PCR

Total RNA was extracted from cells and tissues using the TRNzol Universal Reagent (Tiangen #DP4-02). For cDNA synthesis, we employed the PrimeScript RT reagent Kit with gDNA Eraser (TAKARA #RR047A) following the manufacturer’s protocol. Quantitative PCR was carried out using SuperReal PreMix Plus (SYBR Green) (Tiangen #DP4-02). RP49 served as the endogenous reference gene for normalization across samples. All experiments included three independent biological replicates to ensure reproducibility. The primers used in this experiment are RP49-F: CGGTTACGGATCGAACAAGC; RP49-R: CTTGCGCTTCTTGGAGGAGA; Mettl5-F: CGGTTTCTGGAGGTGGC; Mettl5-R: GCTGGCGTCGATGTTGTAC..

Imaging and analysis of axon volume

We quantified small LNvs axon volume using ImageJ’s Object Counter 3D plugin to measure pixel counts from raw images. Image processing involved applying a standardized threshold that clearly visualized intact axons in control samples. For consistent measurements, we specifically analyzed the axon span between the first axonal bifurcation and the terminal tip, as indicated by the white rectangular markers in Figure 8.

Sample collection and library construction for Ribo-seq and RNAseq

For sample collection, w1118 and Mettl51bp Drosophila were harvested at ZT15, immediately transferred to centrifuge tubes, and flash-frozen in liquid nitrogen. Fly heads were subsequently separated and collected while frozen, with three biological replicates prepared for each genotype. Each sample was equally divided for parallel RNA-seq and Ribo-seq analyses. For Ribo-seq library preparation, samples were lysed in buffer containing 50 mg/mL cycloheximide (Novogen, China) to preserve ribosome positioning, followed by RNase I digestion to generate ribosome-protected fragments (RPFs). Monosomes were isolated using MicroSpin S-400 HR size-exclusion chromatography, followed by rRNA depletion and PAGE purification to select 20-38 nt RPFs. Purified fragments underwent end repair, adapter ligation, reverse transcription, and PCR amplification before Illumina PE150 sequencing. For RNA-seq, total RNA was extracted using TRIzol reagent, with cDNA libraries prepared and sequenced using Illumina PE150 by Novogen. Raw sequencing data in FASTQ format were processed to remove adapter sequences, reads containing N bases, and low-quality reads, while simultaneously calculating Q20/Q30 scores and GC content to generate clean reads for downstream analysis.

RNA-seq analysis

We performed genome alignment and transcriptome analysis using the following pipeline: First, we built a Hisat2 index (v2.0.5) for the Drosophila melanogaster reference genome (dm6 assembly). Clean paired-end reads were then aligned to this reference using Hisat2 (v2.0.5) (Kim et al., 2019). The resulting alignments were processed using StringTie (v1.3.3b) (Pertea et al., 2015) for reference-based transcript assembly. For gene-level quantification, we used FeatureCounts (v1.5.0) (Liao et al., 2014) to count reads mapped to each annotated gene. Differential expression analysis between w1118 and Mettl51bp was conducted with DESeq2 (v1.20.0) (Pertea et al., 2015), with genes meeting both criteria (adjusted p-value < 0.05 and absolute log2 fold change ≥ 1) considered statistically significant.

Ribo-seq analysis

We implemented a comprehensive Ribo-seq analysis pipeline beginning with quality filtering using Bowtie (Liao et al., 2014) to remove reads aligning to non-coding RNAs (rRNA, tRNA, snoRNA, snRNA from FlyBase Release 6.13) with a 2-mismatch allowance (-v 2). The remaining ribosome-protected fragments (RPFs) were mapped to the Drosophila melanogaster genome (FlyBase Release 6.13) using STAR (v2.7.3a) (Dobin et al., 2013), followed by transcript-level alignment to protein-coding sequences using Bowtie (v1.2.2) (Langmead et al., 2009) with parameters "-a -v 2". CDS-aligned RPFs were quantified using featureCounts (Subread v1.6.3) and normalized as RPKM. Differential expression analysis was performed with DESeq2 (v1.14.1) (Love et al., 2014) using thresholds of |log2FC| ≥ 0.265 and padj < 0.05, while translation efficiency differences were assessed using RiboDiff. Data quality was verified through riboWaltz (v1.1.0) (Lauria et al., 2018) for 3-nt periodicity and reading frame analysis, with Ribocode (Xiao et al., 2018) employed for P-site positioning and uORF motif analysis. Functional enrichment analyses (GO, KEGG, GSEA) were conducted using clusterProfiler (v4.5.2.002) (Wu et al., 2021), with results visualized through ggplot2-generated plots and GO term simplification performed using SimplifyEnrichment (Gu and Hübschmann, 2022).

Immunofluorescence experiments

We performed immunofluorescence staining on 7-15 days old adult flies (unless otherwise specified). Flies were anesthetized with CO₂ and dissected in ice-cold 0.03% PBST (1× PBS with 0.03% Triton X-100; Sigma, T9284). Samples were fixed in 2% paraformaldehyde (PFA) for 55 minutes at room temperature (RT), followed by four 15-minute washes in 0.03% PBST at RT. After blocking overnight at 4°C in 10% Normal Goat Serum (NGS; in 1× PBS with 2% Triton), samples were incubated with primary antibodies for 24 hours at 4°C. Primary antibodies included rat anti-Elav (DSHB, 9F8A9; 1:200) and mouse anti-Repo (DSHB, 8D12; 1:200), diluted in antibody buffer (1.25% PBST, 1% NGS). Following four 15-minute washes in 1× PBS with 1% Triton at RT, samples were incubated overnight at 4°C with secondary antibodies: Alexa Fluor™ 568 (Thermo Fisher, A11004; 1:200) and Alexa Fluor™ 647 (Thermo Fisher, A21247; 1:200). After four additional 15-minute washes in 1× PBS with 1% Triton at RT, samples were mounted using DAPI-containing antifade mounting medium (Solarbio, S2110).

Images were acquired using a Leica SP8 confocal microscope with LAS X software, applying auto Z-brightness correction when needed for signal uniformity. Images were processed in Adobe Photoshop CS6 and figures assembled in Adobe Illustrator 2020. PDF immunofluorescence followed the same protocol. All experiments included ≥3 biological replicates, each containing ≥10 flies. Fluorescence intensity was quantified using ImageJ.

Western Blotting

Protein samples were prepared by homogenizing approximately 30 fly heads in RIPA lysis buffer (150 mM NaCl, 1.0% NP-40, 0.5% sodium deoxycholate, 0.1% SDS, 50 mM Tris-HCl, pH 8.0) supplemented with protease (CW2200S) and phosphatase (CW2383S) inhibitor cocktails according to manufacturer specifications. Lysates were mixed with 2× SDS loading buffer, boiled at 100°C for 5 min, and immediately cooled on ice.

For immunoblotting, membranes were probed with rabbit anti-PER primary antibody (1:5000 dilution; kindly provided by Dr. Jeffrey Price’s laboratory, University of Missouri-Kansas City) overnight at 4°C, followed by incubation with HRP-conjugated goat anti-rabbit IgG secondary antibody (1:1500; ABclonal, AS014) for 4 hr at room temperature. Protein signals were detected using ECL substrate (ABclonal, RM00021P) and imaged with an Amersham ImageQuant 800 system (GE Healthcare). Band intensities were quantified using ImageJ software, with three biological replicates performed for statistical reliability.

LC-MS/MS analysis of m6A levels

Total RNA was isolated from cells and tissues using TRNzol Universal Reagent (Tiangen, #DP4-02). For 18S rRNA purification, we separated total RNA by polyacrylamide gel electrophoresis followed by gel extraction. Our analysis utilized ribonucleoside standards including adenosine (rA) and N6-methyladenosine (N6 mA), with a mobile phase consisting of methanol:ddH2O (v/v).

For each biological replicate, 1 μg of total RNA or purified 18S rRNA was digested to single nucleosides using Nucleoside Digestion Mix (NEB, #M0649). Proteins were precipitated by adding a 4:1 methanol:digest ratio and incubating at -20°C for 2 hours. Quantitative analysis was performed using multiple reaction monitoring (MRM) with the following transitions: 268.10275→136.0621 (rA), 282.11835→150.0774 (N6 mA).

Negative geotaxis RING assay

Climbing assays were performed following a 12-hour recovery period after CO2 anesthesia, using 200 flies per genotype distributed across 10 vials (20 flies/vial) marked at 90 mm height. Each assay consisted of three trials separated by 15-minute intervals, initiated by sharply tapping vials three times to induce negative geotaxis, with the number of flies reaching the 90 mm mark within 10 seconds recorded per trial. Five groups were tested per genotype (15 trials total). All assays were video recorded under standardized conditions with the camera positioned 20 cm from vials and uniform backlighting provided by a white open-faced box to ensure consistent imaging quality.

eLife Assessment

The authors present useful findings demonstrating that the RNA modification enzyme Mettl5 regulates sleep in Drosophila. Through transcriptome- and proteome-wide analyses, the authors identified downstream targets affected in heterozygous mutants and proposed that Mettl5 regulates the translation and degradation of clock genes to maintain normal sleep function. Through additional analyses, the authors provided solid evidence that Mettl5 regulates translation and degradation of clock genes to maintain normal sleep cycle. The mechanistic details of Mettl5 is unclear and requires further support.

Reviewer #1 (Public review):

Here, the authors attempted to test whether the function of Mettl5 in sleep regulation was conserved in drosophila, and if so, by which molecular mechanisms. To do so they performed sleep analysis, as well as RNA-seq and ribo-seq in order to identify the downstream targets. They found that the loss of one copy of Mettl5 affects sleep, and that its catalytic activity is important for this function. Transcriptional and proteomic analyses show that multiple pathways were altered, including the clock signaling pathway and the proteasome. Based on these changes the authors propose that Mettl5 modulate sleep through regulation of the clock genes, both at the level of their production and degradation, possibly by altering the usage of Aspartate codon.

Comments on revised version:

The authors satisfactorily addressed my comments, even though the precise mechanism by which Mettl5 regulates translation of clock genes remains to be firmly demonstrated.

Reviewer #3 (Public review):

Xiaoyu Wu and colleagues examined a potential role in sleep of a Drosophila ribosomal RNA methyltransferase, mettl5. Based on sleep defects reported in CRISPR generated mutants, the authors performed both RNA-seq and Ribo-seq analyses of head tissue from mutants and compared to control animals collected at the same time point. A major conclusion was that the mutant showed altered expression of circadian clock genes, and that the altered expression of the period gene in particular accounted for the sleep defect reported in the mettl5 mutant. In this revision, the authors have added a more thorough analysis of clock gene expression and show that PER protein levels are increased relative to wild type animals a specific times of day, indicating increased stability of the protein. Given that PER inhibits its own transcription, the per RNA is low in the mutants. Efforts toward a more detailed understanding of how clock gene expression was altered in the mutants, as well as other clarification of sleep phenotypes throughout is appreciated. As noted above, a strength of this work is its relevance to a human developmental disorder as well as the transcriptomic and ribosomal profiling of the mutant. However, there still remain some minor weaknesses in the manuscript. This reviewer is not in agreement with the interpretation of the epigenetic experiments. Specifically, co-expression of Clk[jrk] or per[01] with the mettl5 mutant recovered the nighttime sleep phenotype, but was additive to the daytime sleep phenotype such that double mutants showed higher sleep. This effect should be acknowledged and discussed. Overall, this is an interesting paper that indicates a molecular link between mettl5 and the circadian clock in regulation of sleep.

Author Response:

The following is the authors’ response to the previous reviews

Public Reviews:

Reviewer #1 (Public review):

Here, the authors attempted to test whether the function of Mettl5 in sleep regulation was conserved in drosophila, and if so, by which molecular mechanisms. To do so they performed sleep analysis, as well as RNA-seq and ribo-seq in order to identify the downstream targets. They found that the loss of one copy of Mettl5 affects sleep, and that its catalytic activity is important for this function. Transcriptional and proteomic analyses show that multiple pathways were altered, including the clock signaling pathway and the proteasome. Based on these changes the authors propose that Mettl5 modulate sleep through regulation of the clock genes, both at the level of their production and degradation, possibly by altering the usage of Aspartate codon.

Comments on revised version:

The authors satisfactorily addressed my comments, even though the precise mechanism by which Mettl5 regulates translation of clock genes remains to be firmly demonstrated.

Reviewer #3 (Public review):

Xiaoyu Wu and colleagues examined a potential role in sleep of a Drosophila ribosomal RNA methyltransferase, mettl5. Based on sleep defects reported in CRISPR generated mutants, the authors performed both RNA-seq and Ribo-seq analyses of head tissue from mutants and compared to control animals collected at the same time point. A major conclusion was that the mutant showed altered expression of circadian clock genes, and that the altered expression of the period gene in particular accounted for the sleep defect reported in the mettl5 mutant. In this revision, the authors have added a more thorough analysis of clock gene expression and show that PER protein levels are increased relative to wild type animals a specific times of day, indicating increased stability of the protein. Given that PER inhibits its own transcription, the per RNA is low in the mutants. Efforts toward a more detailed understanding of how clock gene expression was altered in the mutants, as well as other clarification of sleep phenotypes throughout is appreciated. As noted above, a strength of this work is its relevance to a human developmental disorder as well as the transcriptomic and ribosomal profiling of the mutant. However, there still remain some minor weaknesses in the manuscript. This reviewer is not in agreement with the interpretation of the epigenetic experiments. Specifically, co-expression of Clk[jrk] or per [01] with the mettl5 mutant recovered the nighttime sleep phenotype, but was additive to the daytime sleep phenotype such that double mutants showed higher sleep. This effect should be acknowledged and discussed. Overall, this is an interesting paper that indicates a molecular link between mettl5 and the circadian clock in regulation of sleep.

Recommendations for the authors:

Reviewer #3 (Recommendations for the authors):

The authors misunderstood my original comment for Fig 1A. Please provide an explanation for the significance of the boxed region. There is little or no detail in the legend to help guide the reader.

The information has been added to the figure legends for Figure 1A.

Efforts toward improving analysis of circadian genes as well as sleep phenotypes (sleep onset time, rebound, etc) is much appreciated, thank you. However, Figure S1H and G panel labels are mixed up; please label in the order that they appear and that they correspond to the main text. Why is Figure S1H labeled "ZT 14"?

Sleep latency is defined as the time from preparing to sleep to actually falling asleep. In this study, it specifically refers to the time taken for each individual fly to reach the sleep phenotype (i.e., 25 minutes of continuous sleep). We noted that this label was misleading, as the actual time to reach the sleep phenotype varied among individual flies. Therefore, in the revised figures, we have removed the ZT14 label. In addition, we have corrected the labeling of Figures S1G and S1H to ensure they appear in the correct order and correspond accurately to the descriptions in the main text.

Unfortunately, based on Fig S1A-C, I am not convinced that mettl5 localizes to neurons, as there are no cells that show double labelling. This figure does not support the statement: "we found expression in both neurons (colocalizing with ELAV staining: Figure S1A-C) (lines 91-92), and "Mettl5-Gal4 is expressed in distinct neurons and glia that appear crucial for sleep regulation." (line 297). What "distinct" sleep related neurons were labeled? The staining in Fig S1A shows a different distribution from that in Fig S1D, and so it's possible this was a technical issue. Is there a better example?

Thank you for your careful review and valuable comments. We agree that the colocalization of METTL5 with the neuronal marker ELAV is relatively sparse. However, as indicated by the arrows in Fig S1A–C, we did observe a few cells showing clear double labeling. These examples support the presence of METTL5 expression in neurons, albeit at a low frequency.

In Figure 4G-H, please indicate the time of day of tissue collection.

In Figure 4G-H, the tissue was collected at ZT0. We have now indicated this time point in the figure and legend to clarify the experimental timing.

As noted in the public comment, I remain in disagreement with the assessment that "the double mutant showed the similar phenotype as downstream genes". The striking significant increase in daytime sleep in the double mutants remains unexplained. No further experiments are necessary, but this should be acknowledged in the text. Instead of an epistatic effect, given that overall sleep is high in the double mutants, another possible explanation is that the flies are sick and so are less active and sleeping more.

Thank you for your suggestion. This has been acknowledged in the text. “Genetic epistasis experiments further supported this model, with clock gene mutants modified Mettl5 mutant phenotypes that suggesting both Clock and Per downstream of Mettl5 (Figure 4I-N, Table 1). Secondary effect may exist for the significant increase in daytime sleep in the double mutants.”

Version 3

Abstract

Sleep plays a critical role in animal physiology, primarily governed by the brain, and its disruption is prevalent in various brain disorders. Mettl5 is associated with intellectual disability (ID), which often includes sleep disturbances. However, the mechanism underlying these sleep disruptions in ID remains poorly understood. In this study, we investigated the sleep phenotypes resulting from Drosophila Mettl5 mutations. Rescue experiments revealed that Mettl5 functions predominantly within neurons and glia marked by Mettl5-Gal4 to regulate sleep. Previous work established that Mettl5 forms a complex with Trmt112 to influence rRNA methylation. Notably, a mutation in Trmt112 mutation recapitulated these sleep disturbances, implicating translational regulation by the Mettl5/Trmt112 complex. Subsequent RNA-seq and Ribo-seq analyses of Mettl51bp mutants uncovered downstream effects, including altered expression of proteasome components and clock genes. Rescue experiments confirmed that the net increase in PERIOD protein underlies the sleep phenotype. This study illuminates the interplay between ribosome function, clock genes, and the proteasome in sleep regulation, highlighting the integrated roles of protein synthesis and degradation. These findings could potentially provide an example for in vivo study of rRNA methylation function, expand our understanding of protein homeostasis in sleep and offer insights into the sleep phenotypes associated with ID.

Introduction

Sleep is essential for animal physiology, and understanding its molecular mechanisms has significant implications for both basic research and clinical applications. Although previous studies have identified key sleep regulators (Du et al., 2021), fundamental questions remain regarding the roles of cellular processes such as protein synthesis and degradation in sleep regulation.

Growing evidence suggests a dynamic interplay between sleep regulation and protein homeostasis. Protein synthesis is particularly active during sleep (Lyons et al., 2023), while Sleep deprivation affects translational initiation (Costa et al., 2019). The proteasome, a critical mediator of ubiquitin-dependent protein degradation, regulates synapse homeostasis by modulating ribosomal components (Costa et al., 2019), thereby influencing synaptic function. Intriguingly, proteasome components exhibit oscillating expression pattern at the transcriptional level in human cells (Desvergne et al., 2016). A recent study in Drosophila showed that mutation in a proteasome component alters sleep patterns (Fernández et al., 2020). Despite these advances, the precise mechanisms by which protein synthesis and degradation influence sleep, and how these processes interact, remain unresolved.

While several factors influencing proteasome assembly and subunit transcription have been identified (Kapetanou et al., 2022), the translational control of proteasome subunits remains poorly understood. Additionally, proteasome inhibition has been shown to alter ribosome function (Galimberti et al., 2016; Costa et al., 2019; Palanca et al., 2014), raising the question of whether the proteasome itself is regulated by the status of the translation machinery. Exploring this relationship could provide key insights into the interplay between proteasome activity and translation.

In Drosophila, Mettl5 exhibits 18S ribosomal RNA m6A methyltransferase activity (Leismann et al., 2020) and interacts with Trmt112 to facilitate this function, a mechanism conserved in human (Leismann et al., 2020; van Tran et al., 2019). Although Mettl5 depletion abolishes m6A modification on 18S rRNA without impairing rRNA maturation, it affects fly orientation behavior. In mammalian cells, Mettl5-mediated 18S rRNA N6-methyladenosine (m6A) modification regulates stem cell fate determination and neural function (Wang et al., 2020). However, it remains unclear whether Mettl5 modulates global or transcript-specific translation profile in vivo, particularly in tissues rather than cell lines. Moreover, while Mettl5 is linked to intellectual disability (ID) with comorbid sleep disturbances, its mechanistic role in these disorders is unknown. Testing whether Mettl5 loss affects behaviors like sleep in Drosophila could clarify this connection.

We discovered that Mettl5 regulates sleep in Drosophila. To investigate the underlying mechanism, we performed RNA-seq and Ribo-seq on Mettl5 mutants, revealing dysregulation of multiple clock genes and proteasome components. This suggests Mettl5 coordinates protein production and degradation, which are crucial for protein homeostasis. Follow-up experiments confirmed that the protein level of Period was upregulated in Mettl5 mutants, contributing to the sleep phenotype. This study highlights that ribosome defects can perturb proteasome function, uncovering a mechanism that couples protein degradation with synthesis. Additionally, by mapping the genome-wide downstream gene profile of an rRNA methylation modifier, this study offers insights into the gene-specific roles of ribosome function.

Results

Mettl5 regulates sleep in Drosophila

Mettl5 contains a predicted N6 adenine specific nucleic acids methyltransferase domain. In order to study the function of this gene in sleep regulation, we generated CRISPR-Cas9 knockout mutants. We created two alleles. Mettl51bp, which deleted 1bp in the coding region results in a truncated version (Figure 1A) and down regulation of Mettl5 at mRNA level (Figure 1B). Another allele Mettl59bp, which results in 3 amino acids deleted version of Mettl5 (Figure 1A) does not cause significant change at mRNA level (Figure 1B). We found that heterozygous Mettl5¹ᵇᵖ mutants exhibited significantly reduced nighttime sleep (Figure 1C-G), particularly during early night (ZT12-ZT16; Figure 1C). Quantitative analysis revealed increased wakefulness at ZT14 in mutants (Figure 1H). ‘awake %’ was used to indicate the percentage of awake fruit fly population at specific time points (e.g., ZT14). quantitative nighttime sleep latency measurements indicated a delayed sleep start in mutants (Figure S1G). In addition, Mettl5 expression was significantly up regulated during the sleep recovery period after mechanical sleep deprivation (Figure 1I-J). Mettl5 mutant displayed significantly increased sleep rebound in 24 hours after sleep deprivation (Figure 1K-L), indicating its effects on sleep homeostasis. Moreover, results of sleep arousal assay at ZT19 indicated that the percentage of aroused flies are significantly more than the control group (Figure 1M). These results demonstrate that the truncated Mettl5 mutation causes sleep deficits, establishing Mettl5 as a novel sleep regulator in Drosophila.

Mettl5 is a regulator of Drosophila sleep.

(A) Diagram illustrating CRISPR-Cas knockout of 1 or 9 bases in the Mettl5 gene. The corresponding protein sequence is listed with the predicted N6 adenine specific nucleic acids methyltransferase domain highlighted in the red box. (B) Relative expression of Mettl5 mRNA in homozygous Mettl51bp and Mettl59bpmutant male flies compared to control flies. (C) Sleep curve throughout the day for Mettl5 mutant male flies (blue) and control flies (black). (D) Total sleep of Mettl5 mutant male flies and control flies in 24 hours. (E) Total sleep of Mettl5 mutant male flies and control flies within day and night, respectively. (F) Sleep bout duration of Mettl5 mutant male flies and control flies. (G) Number of sleep bouts of Mettl5 mutant male flies and control flies. (H) Percentage of awake for Mettl5 mutant flies and control flies. (I) Sleep curve is tracked throughout the entire day prior to sleep deprivation and during the daytime sleep rebound period. (J) Mettl5 mRNA expression level at different time points. W(wake), SD (sleep deprivation), SR (sleep recovery). (K) Sleep curve is tracked throughout the entire day prior to sleep deprivation and during the daytime sleep rebound period in Mettl5 mutant male flies (blue) and control flies (black). (L) Response to sleep deprivation and performance measures in Mettl5 mutants and controls. Black bars represent the amount of sleep lost during the 24-hour sleep deprivation period, blue bars indicate the amount of sleep regained, whereas the red bars indicate the proportion of sleep recovered (right y-axis). (M) Sleep arousal of Mettl51bp male flies and control flies at ZT19. (N) Sleep curve throughout the day for the following genotypes: w1118(black), Mettl51bp/+ (blue), and Mettl5-Gal4, UAS-Mettl5, Mettl51bp/+(pink). (O) Total sleep of the indicated genotypes. (P) Sleep bout duration of the indicated genotypes. (Q) Number of sleep bouts of the indicated genotypes. (R) Percentage of awake for the indicated genotypes. For * stands for p<0.05, ** stands for p<0.01, *** stands for p<0.001, ns stands for not significant. For letter-based annotations, Groups with no significant differences share the same letter, Groups with significant differences are assigned new letters.

To confirm the specificity of the sleep phenotype in Mettl51bp, we performed genetic rescue experiments. Introduction of a single wild-type Mettl5 copy completely rescued both the reduced sleep amount and increased wakefulness at ZT14 (Figure 1N-R), demonstrating that these phenotypes specifically result from Mettl5 deficiency.

The observed expression pattern of Mettl5 further supports its sleep regulatory function. Using Mettl5-Gal4 reporter lines, we found expression in both neurons (colocalizing with ELAV staining; Figure S1A-C) and glial cells (colocalizing with Repo staining; Figure S1D-F). Behavioral analyses revealed additional mutant phenotypes consistent with sleep dysregulation. we tested the climbing ability of Mettl5 mutation which showed an increase (Figure S1H). RNAi knocking down of Mettl5 showed a consistent phenotype of down regulated sleep amount during the night time (Figure S1I-L).

Mettl5 regulates Drosophila sleep through its methyltransferase activity

Previous studies established that Mettl5 interacts with Trmt112 to regulate 18S rRNA m6A modification (Leismann et al., 2020; van Tran et al., 2019). Consistent with these findings, our LC-MS/MS analysis revealed significantly reduced m⁶A levels in both total RNA and 18S rRNA from heterozygous Mettl51bp (Figure 2A-B). To determine whether Mettl5’s sleep regulatory function depends on its methyltransferase activity, we performed two key experiments. First, we examined the phenotype of Trmt112 knockdown. We found that the Mettl5-Gal4 driven Trmt112 RNAi resulted in a similar phenotype to Mettl51bp (Figure 2C-G), suggesting that Mettl5 regulates sleep through its 18S rRNA m6A modification activity. The efficiency of the Trmt112 RNAi line was tested in previous study (López-Varea et al., 2021). More important, we performed a rescue experiment with a mutated form of Mettl5 that lacks the NPPF amino acids required for its methyltransferase activity (Figure2A) (Iyer et al., 2016). We observed that this mutant Mettl5 failed to rescue the sleep phenotype caused by Mettl51bp (Figure 2H-L). Together, these results demonstrate that Mettl5’s methyltransferase activity is essential for its role in sleep regulation, likely through its function in 18S rRNA m⁶A modification.

Mettl5 regulation of Drosophila sleep was dependent on its methyltransferase activity.

(A) The m6A level in the total RNA of Mettl5 mutant male flies. (B) The m6A level in the 18S rRNA of Mettl5 mutant male flies. (C) The sleep curve throughout the day shows the sleep pattern of induced Trmt112 RNAi male flies and control flies. (D) Total sleep of induced Trmt112 RNAi male flies and control flies. (E) Sleep bout duration in induced Trmt112 RNAi male flies and control flies. (F) Number of sleep bouts in induced Trmt112 RNAi male flies and control flies. (G) Percentage of awake in Trmt112 RNAi and control flies. (H) Sleep curve throughout the day for Mettl5 mutant male flies, induced Mettl5m over expression male flies and control flies. (I) Total sleep of Mettl5 mutant male flies, induced Mettl5m over expression male flies and control flies. (J) Sleep bout duration in Mettl5 mutant male flies, induced Mettl5 over expression male flies and control flies. (K) Number of sleep bouts in Mettl5 mutant male flies, induced Mettl5 over expression male flies and control flies. (L) Percentage of awake in Mettl5 mutant male flies, induced Mettl5 over expression male flies and control flies. For * stands for p<0.05, ** stands for p<0.01, *** stands for p<0.001, ns stands for not significant. For letter-based annotations, Groups with no significant differences share the same letter, Groups with significant differences are assigned new letters.

RNA-seq and Ribo-seq revealed the downstream gene profile of Mettl5

To better understand the downstream events of Mettl5, we performed RNA-seq and Ribo-seq to assess transcriptomic and translational changes in Mettl51bp. Principal Coordinates Analysis (PCoA) of RNA-seq and Ribo-seq datasets revealed clear separation between mutant and control groups (Figure S2A, B, D, E). High reproducibility was found among biological replicates (Figure S2C, F), confirming data quality. Transcriptome analysis identified 1,053 significantly differentially expressed genes (|log2 (Fold Change) | ≥ 1 & padj< 0.05), comprising 217 upregulated and 836 downregulated transcripts (Figure 3A, B). Parallel ribosome profiling revealed 299 translationally regulated genes (|log2 (Fold Change) | ≥ 0.265 & padj< 0.05), with 149 upregulated and 150 downregulated targets compared to the controls (Figure 3C). Heatmaps of the top 100 differentially expressed genes from both RNA-seq and Ribo-seq analyses were generated according to the rlog transformed values (Figure S3A, B).

RNA-seq and Ribo-seq analysis revealed changes in the gene profile of Mettl51bp.

(A) Venn diagram depicting the number of significant differentially expressed genes revealed by RNA-seq and Ribo-seq. (B) Volcano plot representing the differentially expressed genes identified by RNA-seq. Genes that met the criteria of |log2(fold-change)| ≥ 1 and p.adjust < 0.05 were considered significantly expressed, marked in orange for downregulation and green for upregulation, comparing with the controls. (C) Volcano plot representing the differentially expressed genes identified by Ribo-seq. Candidates that satisfied the criteria of |log 2 (fold-change) | ≥ 0.265 and p.adjust <0.05 were regarded as significantly expressed, marked in red for downregulation and blue for upregulation, respectively. (D-K) Gene set enrichment analysis of differentially expressed genes revealed by RNA-seq (D-G) and Ribo-seq (H-K). All the plots are generated using the KEGG gene set database. The bar chart at the bottom of each panel shows the distribution of target genes for each pathway according to their rank position. Each vertical line represents a gene. Genes on the left show positive correlation with Mettl51bp, while, genes on the right show negative correlation with Mettl51bp. The green line indicates the enrichment score (ES), and NES stands for normalized enrichment score. (L) Distribution of the differentially expressed genes revealed by both RNA-seq and Ribo-seq. (M) Cumulative distribution of TE frequencies among w1118 and Mettl51bp (N, O) GO and KEGG enrichment of significantly changed TE-related genes between w1118and Mettl51bp. The color of the bar indicates the enrichment p.adjust value. (P) KEGG network showing the top 10 pathway and associated genes. The size of the dots represents the number of genes in the pathway.

To explore the biological implication of differentially expressed genes at transcriptional and translational levels, we performed Gene Ontology (GO) enrichment analysis. At the transcriptional level (Figure S4A, C), the most significantly enriched biological processes included cellular response to chemical stimulus, small molecule biosynthetic process, and fatty acid metabolic process (Figure S4A), with supporting gene networks (Figure S4C). Translational-level analysis (Figure S4B, D) revealed prominent enrichment for organic acid and amino acid metabolic processes, lipid catabolism, cellular respiration, and transport mechanisms. Notably, we identified strong associations with circadian regulation, including circadian regulation of gene expression and entrainment of the circadian clock.

The GO results usually contain a long list of enriched terms that have highly redundant information and are difficult to summarize. So, we performed a simplify Enrichment analysis that visualizes the summaries of clusters by word cloud for the GO enrichment result at transcriptional and translational levels. For transcriptionally significant DEGs (Figure S5A), major enriched processes included metabolic and stimulus-response pathways, wing disc and imaginal system development, ion-channel homeostasis and transport, and cytokinesis/cell cycle transitions. Translationally regulated genes additionally showed enrichment for sleep-wake cycles, circadian behavior, and neuronal cell death (Figure S5D). Cellular component analysis highlighted chromosomal structures, membrane-bound complexes, and vesicles (Figure S5B), while molecular functions predominantly involved enzymatic activities, ion/protein binding, receptor activities, and transcription processes (Figure S5C). The strong concordance between RNA-seq and Ribo-seq enrichment profiles (Figure S5E-F) further validated these findings.

Our Gene Set Enrichment Analysis (GSEA) identified pathways enriched in differentially expressed genes at both transcriptional and translational levels. To capture subtle but biologically important changes that might be excluded by stringent statistical thresholds, we analyzed gene sets ranked by log2 fold-change without p-value filtering. The GSEA-GO and GSEA-KEGG results, sorted by Normalized Enrichment Score (NES), revealed several key pathways (Figure S6A-D). Notably, the proteasome pathway showed significant suppression at both transcriptional and translational levels (P.adjust<0.05) (Figure 3E, I). Transcriptional-level analysis highlighted disruptions in arginine and proline metabolism and insect hormone biosynthesis (P.adjust<0.05) (Figure 3D, G). While, while translational changes predominantly affected fatty acid degradation and phosphonate/phosphinate metabolism (P.adjust<0.05) (Figure 3H-K).

To better understand the correlation between transcription and translation levels, we performed correlation analysis of differentially expressed genes based on fold change of DEGs (Figure 3L). 977 genes showed opposing trends between transcription and translation, while only 24 genes exhibited concordant changes. We identified 240 translation-specific and 3,309 transcription-specific DEGs, with each cluster undergoing separate enrichment analysis (Figure S7A-H). Notably, core clock genes (Clk, tim, etc.) displayed significant changes at both levels (Figure 3L).

To isolate translational effects, we calculated translation efficiency (TE) differences between Mettl51bp and w1118controls (Figure 3M). Among 1,204 genes with significantly altered TE, GO enrichment highlighted amino acid metabolism and small molecule biosynthesis (Figure 3N), while KEGG analysis emphasized glycine/serine/threonine metabolism and one-carbon pool by folate pathways (Figure 3O, P).

Ribo-seq revealed the Mettl51bp led to changes of some global translation features

Using Ribo-seq, we compared global translation features between the w1118 and Mettl51b groups. The TE (translation efficiency) correlation coefficients among the three biological replicates of each genotype ranged from 0.87 to 0.93, demonstrating high reproducibility within genotypes (Figure S8A-F). We found that the length of ribosome-protected RNA fragments (RPFs) was approximately 28nt (Figure S10A). The RPFs exhibited a significant 3-nt periodicity. Metagene analysis of individual 28nt reads revealed the distribution of RPFs across the gene locus in w1118 and Mettl51bp, respectively (Figure S8, S9). The starting point of translation is 12nt upstream of start-codon and gradually disappears 15nt from stop-codon (Figure S8G-L). RPFs on the metagene plot distribution around the translation start and translation stop site (Figure S9A-F), different coding frames on CDS, 3’-UTR and 5’-UTR, respectively in two groups of samples all showed the periodicity (Figure S9G-L). As expected, all plots show an enrichment of P-sites in the first frame on the coding sequence but not the UTRs, in accord with ribosome protected fragments from protein coding mRNAs (Figure S9G-L).

Analysis of open reading frame (ORF) types revealed alterations in their distribution. Compared to w1118, Mettl51bp exhibited an increased proportion of overlapping downstream ORFs (dORFs) and fewer non-overlapping dORFs (Figure S10B). In both groups, translated upstream ORFs (uORFs) were significantly shorter than untranslated uORFs (P < 0.05) (Figure S10C). Additionally, Mettl51bp displayed reduced read counts for both dORFs and uORFs (Figure S10D-E). Based on translation potential, uORFs were classified as either translated or untranslated, and motif analysis was performed separately for each category (Figure S10F-G).

Mutation of Mettl5 altered codon preference

Comparison of codon occupancy (A-site) between the two groups revealed that Mettl51bp preferentially used GAC and GAU, whereas w1118 favored UCC (Figure S10H). This trend was further supported by the cumulative frequency distribution of these codons (Figure S10I-K). Since GAC and GAU both encode aspartate (Asp), we analyzed Asp amino acid occupancy. Intriguingly, Asp was significantly enriched in w1118 during translation (Figure S10L), suggesting that Mettl51bp may exhibit altered translational regulation (see Discussion for details).

Metagene analysis of RPFs revealed distinct translation patterns between Mettl51bp and w1118. The coding sequence (CDS) and flanking regions were segmented into 100 equal bins, and average RPF density was computed for each bin. The resulting plots illustrate differences in ribosome occupancy between Mettl51bp and w1118 along the CDS (Figure S10M), near the translation start site (Figure S10N) and around the translation termination site (Figure S10O). Notably, Mettl51bp and w1118 exhibited divergent ribosome occupancy patterns, particularly along the CDS and near the start codon (Figure S10M, N), suggesting potential differences in translation dynamics and initiation efficiency.

Mettl5 regulates the clock gene regulatory loop

Our findings demonstrate that Mettl51bp disrupts the core clock gene regulatory loop controlling circadian rhythm. We observed significant alterations in both transcriptional and translational levels of multiple clock genes, with Cry and Clock showing upregulation while Tim, per, vri, and pdp1 were downregulated at both levels (Figure 4A-B). Notably, per, vri, and pdp1 exhibited particularly pronounced downregulation in translation efficiency (Figure 4C). These changes occurred without affecting clock neuron morphology at different time points (Figure S11).

Clock genes expression mediated the sleep phenotype caused by Mettl5 mutation.

(A-C) Fold changes in clock genes with significant expression level differences between w1118 and Mettl51bp were observed in RNA-seq, Ribo-seq, and translation efficiency analyses. (D) The gene expression levels of per at 4 different timepoints of w1118 and Mettl51bp. (E) Western blot analysis of PER protein and Tubulin protein at 4 different timepoints, (F) along with their ratio quantification. (G) Representative image of PER protein immunofluorescence staining at ZT0 in the small ventral lateral neurons (small LNvs). (H) Statistical analysis of the immunofluorescence intensity for PER in small LNvs. (I) Sleep curve throughout the day for Mettl51bp, per01, double mutant and control flies. (J) Total sleep for Mettl51bp, per01, double mutant and control flies. (K) Percentage of awake time in Mettl51bp flies, partially rescued by double mutant flies. (L) Sleep curve throughout the day for Mettl51bp, per01, double mutant and control flies. (M) Total sleep for Mettl51bp, per01, double mutant and control flies. (N) Percentage of awake time in Mettl51bp flies, partially rescued by double mutant flies. (O) Fold changes in proteasome subunits with significant expression level differences between w1118 and Mettl51bp were observed in RNA-seq, Ribo-seq, and translation efficiency analyses. For statistical significance, * stands for p<0.05, ** stands for p<0.01, *** stands for p<0.001, ns stands for not significant.

The observed expression patterns revealed an unexpected regulatory relationship: while the canonical clock circuitry positions Per downstream of Clock, our finding that Clock was upregulated while per was downregulated suggests Per may actually function upstream of Clock in Mettl5-mediated regulation. Surprisingly, despite the transcriptional downregulation of per (Figure 4D), we detected increased PER protein levels through both immunostaining and western blot analyses (Figure 4E-H). Detection of the PER protein at different time points indicated that it was increased at both ZT 0 and ZT18 (Figure 4E-F). This apparent contradiction aligns with the observed circadian phenotype, as Mettl5 mutants showed significantly longer period lengths (Table 1), mirroring effects seen when PER stabilization results from reduced kinase activity as previously reported (Philpott et al., 2023). Genetic epistasis experiments further supported this model, with clock gene mutants modified Mettl5 mutant phenotypes that suggesting both Clock and Per downstream of Mettl5 (Figure 4I-N, Table 1). Secondary effect may exist for the significant increase in daytime sleep in the double mutants. Together, these results indicate that Mettl51bp affects circadian regulation through mechanisms that extend beyond transcriptional control, likely involving post-translational regulation of PER protein stability.

To investigate the factors mediating PER protein level changes in Mettl5 mutants, we examined the ubiquitin-proteasome pathway, which plays a well-documented role in Period protein degradation (Grima et al., 2002; Ko et al., 2002; Chiu et al., 2008). Notably, recent evidence indicates that N6-methyladenosine (m6A) regulates the ubiquitin-proteasome system in other biological contexts (Sun et al., 2023). Our integrated analysis of RNA-seq and Ribo-seq data revealed significant downregulation of multiple proteasome pathway components in Mettl51bp mutants at both transcriptional and translational levels (Figures 3E, 3I, 4O), suggesting impaired protein degradation capacity. Based on these findings, we propose a model where Mettl5 regulates circadian function through three interconnected mechanisms: first, by directly modulating proteasome components to control PER protein stability post-translationally; second, by transcriptionally and translationally regulating Per and other clock expression. In this model, Mettl51bp-induced proteasome downregulation leads to PER accumulation, which is responsible for the phenotypes (Figure 5).

A working model illustrating the role of Mettl5 in Drosophila sleep was presented.

Mettl51bp alters axon complexity

Previous studies have established a strong correlation between sleep homeostasis and synaptic complexity (Bushey et al., 2011). Additionally, synaptogenesis has been shown to enhance proteasome activity in axons (Costa et al., 2019). Given our observations of impaired sleep rebound following deprivation and altered expression of proteasome subunits in Mettl51bp mutants, we sought to examine potential effects on synaptic complexity.

To assess synaptic complexity, we adapted an established quantification method (Bushey et al., 2011) using Syt-GFP, a marker that colocalizes with endogenous synaptic vesicles, to visualize presynaptic morphology changes. Control experiments with UAS-Fmr1 and Fmr1 mutations successfully replicated the expected decrease and increase in syt-eGFP signal, respectively (Figure 6A-D). Strikingly, Mettl51bp mutants exhibited significantly increased syt-eGFP fluorescence in presynaptic terminals (Figure 6E-M), indicating altered synaptic complexity.

The axon complexity was found to be affected by Mettl51bp

(A–C) Representative confocal micrographs of small ventral lateral neuron (s-LNv) axonal terminals across different genotypes. Presynaptic structures were visualized using syt-eGFP (green), which colocalizes with endogenous synaptic vesicles: (A) control (B) Fmr1 overexpression and (C) Fmr1 null mutant. (D) Quantification of s-LNv axonal terminal volumes corresponding to genotypes in (A–C). (E–H) Representative images of axonal terminal morphology in control flies at four time points: ZT0, ZT6, ZT12, and ZT18. (I–L) Representative images of axonal terminal morphology in Mettl5 mutant flies at ZT0, ZT6, ZT12, and ZT18. (M) Quantitative comparison of axonal terminal volumes between control and Mettl5 mutant flies at different time points. Scale bar: 50 μm. For statistical significance, * stands for p<0.05, ** stands for p<0.01, *** stands for p<0.001, ns stands for not significant.

Discussion

Our study reveals that Mettl5, a known rRNA methyltransferase, modulates sleep through its RNA methylation activity. Through integrated RNA-seq and Ribo-seq analyses of Mettl51bpmutants, we identified Mettl5’s downstream targets at both transcriptional and translational levels. Further investigation demonstrated that Mettl5 influences sleep regulation by affecting two key pathways: the circadian clock gene network and the proteasome system. These findings provide novel mechanistic insights into sleep control, highlighting the coordinated role of protein synthesis and degradation in this process. Notably, our Ribo-seq analysis revealed that Mettl51bp alters fundamental translation features, including uORF translation efficiency and codon preference, suggesting rRNA methylation plays a regulatory role in these processes.

This discovery has important clinical implications, as METTL5, the human ortholog of Mettl5, is associated with intellectual disability (ID) when mutated (Richard et al., 2019). Our work expands the understanding of Mettl5’s molecular function and may inform potential therapeutic strategies for ID. The clinical relevance of our findings is underscored by reports of sleep disturbances, particularly reduced sleep duration, in ID patients - a phenotype that parallels our observations in Mettl51bp mutants. The mechanistic framework established in this study could explain these clinical sleep abnormalities. However, further validation in vertebrate models is needed to determine whether this regulatory mechanism is evolutionarily conserved and applicable to human sleep disorders.

As shown in Table 1, the Mettl51bp/+ mutant exhibits a robust long-period phenotype, with circadian rhythms significantly extended to 28.3 ± 0.4 hours compared to the wild-type’s 23.9 ± 0.05 hours. This prolonged period perfectly aligns with the observed behavioral phenotypes, including delayed nighttime sleep onset, later daytime waking, and the overall shift in sleep profile. This is indeed quite similar to previous report on Period3 variant (Zhang et al., 2016). We think that the prolonged circadian period contributes to the observed sleep phenotype. However, since total sleep time was significantly reduced in the mutant, we cannot attribute the phenotype solely to period lengthening. Furthermore, our 24-hour PER expression analysis in Mettl5 mutants revealed elevated PER protein levels at ZT1 and ZT18, while ZT6 and ZT12 showed no significant changes, with no apparent phase shift. These findings collectively suggest that the phenotype primarily results from PER protein stabilization and accumulation.

We found that Mettl5 heterozygotes showed significant reductions in total RNA and 18S rRNA methylation levels, contrasting with mouse studies where heterozygous knockouts maintained normal 18S rRNA m6A methylation (Sepich-Poore et al., 2022). This discrepancy may stem from either fundamental differences in rRNA methylation regulation between Drosophila and mice, or distinct biological consequences of knockdown versus knockout approaches, as complete gene elimination often triggers compensatory mechanisms (Teng et al., 2013; Rossi et al., 2015; Vu et al., 2015; Ma et al., 2019; El-Brolosy et al., 2019).

Our study uncovers a previously unrecognized connection between circadian clock genes and proteasome function. While previous work demonstrated that the circadian clock rhythmically regulates proteasome components in Drosophila fat bodies under dietary restriction (Hwangbo et al., 2023), we now show that Mettl5 modulates clock protein synthesis and degradation in clock neurons by influencing proteasome activity. This regulation likely occurs through Mettl5-mediated ribosomal methylation in clock neurons, which impacts the proteasome degradation pathway (Costa et al., 2019), ultimately altering clock protein dynamics.

Mettl5 represents a novel integrator of rRNA methylation and proteasome function, providing a mechanism to balance protein synthesis and degradation. Interestingly, we observed specific effects on both transcriptional and translational outputs, with particular proteasome subunits showing differential regulation. This specificity may arise from either selective translational control by Mettl5 or additional layers of regulation through protein-protein interactions. These findings suggest an intricate regulatory network coordinating these processes, though further studies are needed to elucidate the underlying mechanisms.

Our results reveal complex relationships between clock genes and sleep regulation. While cyc loss-of-function alleles show enhanced sleep rebound in females (Shaw et al., 2002), Mettl51bpmutants with elevated PER protein exhibit reduced rebound. Notably, Mettl5’s regulatory effects display tissue specificity - although absent in canonical clock neurons, Mettl5-Gal4 is expressed in distinct neurons and glia that appear crucial for sleep regulation. This expression pattern, combined with Clock’s broader distribution (Patop et al., 2023), suggests specialized circuits for sleep homeostasis that warrant further investigation.

Evidence indicates that sleep functions in development, metabolism and neuronal plasticity (Anafi et al., 2019). Sleep exerts effects on neuronal plasticity by modifying synapses. The synaptic homeostasis theory proposes that sleep has a role in downscale synaptic strength (Tononi and Cirelli, 2006). Indeed, synapse markers progressively decrease during sleep in both mammals and Drosophila brains (López-Varea et al., 2021; Vyazovskiy et al., 2008; Gilestro et al., 2009; Liu et al., 2010) More importantly, evidence indicates that sleep need and protein levels in synapse are tightly linked. Studies of the presynaptic active zone have shown that synaptic plasticity regulates sleep homeostasis (Huang et al., 2020). Genome-wide proteomic studies of synapse in mouse brain indicate that the synaptic proteins peak around dusk and dawn (Noya et al., 2019; Brüning et al., 2019). Sleep deprivation experiments have demonstrated that the sleep drive is significantly dependent on the cycling of proteins and phosphoproteins in synapses, in contrast to mRNAs (Noya et al., 2019; Brüning et al., 2019). These results suggest that the gene expression regulation at the protein level is crucial for sleep. Consistently, we detected changes in the quantity of axons in Mettl51bp, suggesting a possible regulation of neural circuits by Mettl5.

The mechanism by which METTL5 regulates translation warrants further investigation. Previous studies have demonstrated that METTL5 influences translation (Rong et al., 2020; Peng et al., 2022), but whether the mechanisms identified here are conserved across other systems remains an intriguing question. In our analysis, we observed increased usage of aspartate (Asp) codons in Mettl5 mutants. Notably, prior work has linked codon usage to PER protein function—specifically, a codon-optimized version of PER failed to rescue circadian rhythmicity in per mutant flies, unlike the wild-type version (Fu et al., 2016). Further analysis revealed that PER protein levels were elevated in these mutants, suggesting that codon optimization enhances PER expression (Figure 2B in Fu et al., 2016). Strikingly, when we examined the codon-optimized region from Fu et al. (2016), we found that GAC (Asp) was highly enriched, raising the possibility that Mettl5 mutation affects PER protein accumulation by altering GAC codon usage. Additional experiments will be needed to validate this hypothesis. Furthermore, we detected changes in upstream open reading frames (uORFs) in Mettl5 mutants, but their relationship to translational regulation requires further exploration.

Our study demonstrates that ribosomal components can exert gene-specific regulatory functions, building upon previous work showing context-dependent ribosome specialization (Simsek et al., 2017). In Mettl5 mutants, we observed distinct alterations in the translational efficiency of specific genes. These effects could potentially arise through two non-exclusive mechanisms: (1) Mettl5-mediated rRNA modifications may modulate ribosomal binding affinity for particular mRNA sequences, or (2) these modifications might contribute to the formation of specialized ribosome populations that preferentially translate specific subsets of mRNAs. Future studies will be required to distinguish between these possibilities and fully elucidate the underlying molecular mechanisms.

Material and methods

Drosophila Strains

Fly stocks used in this study were maintained under standard culture conditions. We used the w1118 as the control strain. The following flies were obtained from Bloomington Stock Center: w1118 (Bl: 5905), Mettl5-Gal4 (Bl: 19514), Repo-Gal4 (Bl: 7415), nSyb-Gal4 (Bl: 51941), pdf-Gal4 (Bl: 41286), ClkJRK (Bl: 80927), per01(Bl: 80917), Fmr1Δ50M(Bl: 6930), UAS-Trmt112RNAi (VDRC: 101515), UAS-DenMark,UAS-syt.eGFP (Bl: 33064), UAS-GFPstinger (from Yi Rao’s lab), UAS-Fmr1 (Bl: 6931), UAS-Mettl5 (FlyORF: F000760) are from Fly ORF collection. UAS-Mettl5m-3HA (generated in this study).

Mettl51bp and Mettl59bp mutants were generated by the CRISPR/Cas system as described previously (Cheng et al., 2020). The target gRNA was designed with an online tool: http://tools.flycrispr.molbio.wisc.edu/targetFinder/. A target sequence was chosen that has the sequence of 5’- GAGTGGGGTACTGTTCCAACAGG with the PAM sequence in bold. Mutations were verified by genomic PCR and sequencing.

Sleep behavior assays

All sleep assays were conducted in a controlled environment incubator maintained at 25 ± 1°C with 60% ± 5% relative humidity. We maintained a 12:12 light-dark cycle with lights on at ZT0 (06:30) and off at ZT12 (18:30). Fly activity was monitored using the Drosophila Activity Monitoring System (Trikinetics, Waltham, MA). Following a 2-day acclimation period, we recorded locomotor activity at 1-minute intervals for 3 consecutive days. Data analysis was performed using pysolo (Gilestro and Cirelli, 2009), with sleep defined as ≥5 minutes of continuous inactivity.

Mechanical sleep deprivation was performed using the SNAP method to keep flies awake for 12 h overnight (Shaw et al., 2002). Sleep deprivation is applied by mechanical stimuli using a timer-controlled rotating shaker. The intensity of stimuli was 1500 rpm once every minute (2 s/1 min) (Shimizu et al., 2008). Baseline sleep was established from 24-hour recordings prior to deprivation. Sleep loss and recovery were quantified according to previous publication (Cirelli C et al., 2005).

Regarding the ‘awake %’ metric, it indicates that at specific time points (e.g., ZT14), the percentage of awake fruit fly population at that moment. At ZT19, we evaluated arousal thresholds by administering a standardized gentle mechanical stimulus. Responsiveness was determined by monitoring activity for 1 minutes post-stimulation. Flies showing no activity during this window were scored as non-responsive. We calculated arousal percentages from the proportion of flies that awakened in response to stimulation.

Circadian rhythm of individual male flies was measured using the Drosophila Activity Monitoring (DAM) System (Trikinetics). Male flies were loaded individually into glass tubes with a length of 65 mm and an inner diameter of 5 mm. The tubes contained standard cornmeal fly food at one end and were sealed with a cotton stopper at the other end. The flies were entrained to a 12 h light/12 h dark cycle for 3 days and then released to constant darkness for at least six days to measure their rhythmicity.Data analysis is done on a Macintosh computer running the FaasX (Fly activity analysis suite) software.

Statistical analysis

All statistical analyses were performed using GraphPad Prism 5 software. For sleep parameter comparisons, we used nonparametric tests including the two-tailed Mann-Whitney test for pairwise comparisons and one-way ANOVA with Tukey’s post hoc test for multiple comparisons. qPCR data were analyzed using unpaired Student’s t-tests, while lifespan data were evaluated with Log-rank (Mantel-Cox) tests. The specific statistical test used for each experiment is indicated in the corresponding figure. In all analyses, a p-value of less than 0.05 was considered statistically significant.

Quantitative PCR

Total RNA was extracted from cells and tissues using the TRNzol Universal Reagent (Tiangen #DP4-02). For cDNA synthesis, we employed the PrimeScript RT reagent Kit with gDNA Eraser (TAKARA #RR047A) following the manufacturer’s protocol. Quantitative PCR was carried out using SuperReal PreMix Plus (SYBR Green) (Tiangen #DP4-02). RP49 served as the endogenous reference gene for normalization across samples. All experiments included three independent biological replicates to ensure reproducibility. The primers used in this experiment are RP49-F: CGGTTACGGATCGAACAAGC; RP49-R: CTTGCGCTTCTTGGAGGAGA; Mettl5-F: CGGTTTCTGGAGGTGGC; Mettl5-R: GCTGGCGTCGATGTTGTAC..

Imaging and analysis of axon volume

We quantified small LNvs axon volume using ImageJ’s Object Counter 3D plugin to measure pixel counts from raw images. Image processing involved applying a standardized threshold that clearly visualized intact axons in control samples. For consistent measurements, we specifically analyzed the axon span between the first axonal bifurcation and the terminal tip, as indicated by the white rectangular markers in Figure 8.

Sample collection and library construction for Ribo-seq and RNAseq

For sample collection, w1118 and Mettl51bp Drosophila were harvested at ZT15, immediately transferred to centrifuge tubes, and flash-frozen in liquid nitrogen. Fly heads were subsequently separated and collected while frozen, with three biological replicates prepared for each genotype. Each sample was equally divided for parallel RNA-seq and Ribo-seq analyses. For Ribo-seq library preparation, samples were lysed in buffer containing 50 mg/mL cycloheximide (Novogen, China) to preserve ribosome positioning, followed by RNase I digestion to generate ribosome-protected fragments (RPFs). Monosomes were isolated using MicroSpin S-400 HR size-exclusion chromatography, followed by rRNA depletion and PAGE purification to select 20-38 nt RPFs. Purified fragments underwent end repair, adapter ligation, reverse transcription, and PCR amplification before Illumina PE150 sequencing. For RNA-seq, total RNA was extracted using TRIzol reagent, with cDNA libraries prepared and sequenced using Illumina PE150 by Novogen. Raw sequencing data in FASTQ format were processed to remove adapter sequences, reads containing N bases, and low-quality reads, while simultaneously calculating Q20/Q30 scores and GC content to generate clean reads for downstream analysis.

RNA-seq analysis

We performed genome alignment and transcriptome analysis using the following pipeline: First, we built a Hisat2 index (v2.0.5) for the Drosophila melanogaster reference genome (dm6 assembly). Clean paired-end reads were then aligned to this reference using Hisat2 (v2.0.5) (Kim et al., 2019). The resulting alignments were processed using StringTie (v1.3.3b) (Pertea et al., 2015) for reference-based transcript assembly. For gene-level quantification, we used FeatureCounts (v1.5.0) (Liao et al., 2014) to count reads mapped to each annotated gene. Differential expression analysis between w1118and Mettl51bp was conducted with DESeq2 (v1.20.0) (Pertea et al., 2015), with genes meeting both criteria (adjusted p-value < 0.05 and absolute log2 fold change ≥ 1) considered statistically significant.

Ribo-seq analysis

We implemented a comprehensive Ribo-seq analysis pipeline beginning with quality filtering using Bowtie (Liao et al., 2014) to remove reads aligning to non-coding RNAs (rRNA, tRNA, snoRNA, snRNA from FlyBase Release 6.13) with a 2-mismatch allowance (-v 2). The remaining ribosome-protected fragments (RPFs) were mapped to the Drosophila melanogaster genome (FlyBase Release 6.13) using STAR (v2.7.3a) (Dobin et al., 2013), followed by transcript-level alignment to protein-coding sequences using Bowtie (v1.2.2) (Langmead et al., 2009) with parameters “-a -v 2”. CDS-aligned RPFs were quantified using featureCounts (Subread v1.6.3) and normalized as RPKM. Differential expression analysis was performed with DESeq2 (v1.14.1) (Love et al., 2014) using thresholds of |log2FC| ≥ 0.265 and padj < 0.05, while translation efficiency differences were assessed using RiboDiff. Data quality was verified through riboWaltz (v1.1.0) (Lauria et al., 2018) for 3-nt periodicity and reading frame analysis, with Ribocode (Xiao et al., 2018) employed for P-site positioning and uORF motif analysis. Functional enrichment analyses (GO, KEGG, GSEA) were conducted using clusterProfiler (v4.5.2.002) (Wu et al., 2021), with results visualized through ggplot2-generated plots and GO term simplification performed using SimplifyEnrichment (Gu and Hübschmann, 2022).

Immunofluorescence experiments

We performed immunofluorescence staining on 7-15 days old adult flies (unless otherwise specified). Flies were anesthetized with CO₂ and dissected in ice-cold 0.03% PBST (1× PBS with 0.03% Triton X-100; Sigma, T9284). Samples were fixed in 2% paraformaldehyde (PFA) for 55 minutes at room temperature (RT), followed by four 15-minute washes in 0.03% PBST at RT. After blocking overnight at 4°C in 10% Normal Goat Serum (NGS; in 1× PBS with 2% Triton), samples were incubated with primary antibodies for 24 hours at 4°C. Primary antibodies included rat anti-Elav (DSHB, 9F8A9; 1:200) and mouse anti-Repo (DSHB, 8D12; 1:200), diluted in antibody buffer (1.25% PBST, 1% NGS). Following four 15-minute washes in 1× PBS with 1% Triton at RT, samples were incubated overnight at 4°C with secondary antibodies: Alexa Fluor™ 568 (Thermo Fisher, A11004; 1:200) and Alexa Fluor™ 647 (Thermo Fisher, A21247; 1:200). After four additional 15-minute washes in 1× PBS with 1% Triton at RT, samples were mounted using DAPI-containing antifade mounting medium (Solarbio, S2110).

Images were acquired using a Leica SP8 confocal microscope with LAS X software, applying auto Z-brightness correction when needed for signal uniformity. Images were processed in Adobe Photoshop CS6 and figures assembled in Adobe Illustrator 2020. PDF immunofluorescence followed the same protocol. All experiments included ≥3 biological replicates, each containing ≥10 flies. Fluorescence intensity was quantified using ImageJ.

Western Blotting

Protein samples were prepared by homogenizing approximately 30 fly heads in RIPA lysis buffer (150 mM NaCl, 1.0% NP-40, 0.5% sodium deoxycholate, 0.1% SDS, 50 mM Tris-HCl, pH 8.0) supplemented with protease (CW2200S) and phosphatase (CW2383S) inhibitor cocktails according to manufacturer specifications. Lysates were mixed with 2× SDS loading buffer, boiled at 100°C for 5 min, and immediately cooled on ice.

For immunoblotting, membranes were probed with rabbit anti-PER primary antibody (1:5000 dilution; kindly provided by Dr. Jeffrey Price’s laboratory, University of Missouri-Kansas City) overnight at 4°C, followed by incubation with HRP-conjugated goat anti-rabbit IgG secondary antibody (1:1500; ABclonal, AS014) for 4 hr at room temperature. Protein signals were detected using ECL substrate (ABclonal, RM00021P) and imaged with an Amersham ImageQuant 800 system (GE Healthcare). Band intensities were quantified using ImageJ software, with three biological replicates performed for statistical reliability.

LC-MS/MS analysis of m6A levels

Total RNA was isolated from cells and tissues using TRNzol Universal Reagent (Tiangen, #DP4-02). For 18S rRNA purification, we separated total RNA by polyacrylamide gel electrophoresis followed by gel extraction. Our analysis utilized ribonucleoside standards including adenosine (rA) and N6-methyladenosine (N6 mA), with a mobile phase consisting of methanol:ddH2O (v/v).

For each biological replicate, 1 μg of total RNA or purified 18S rRNA was digested to single nucleosides using Nucleoside Digestion Mix (NEB, #M0649). Proteins were precipitated by adding a 4:1 methanol:digest ratio and incubating at -20°C for 2 hours. Quantitative analysis was performed using multiple reaction monitoring (MRM) with the following transitions: 268.10275→136.0621 (rA), 282.11835→150.0774 (N6 mA).

Negative geotaxis RING assay

Climbing assays were performed following a 12-hour recovery period after CO2 anesthesia, using 200 flies per genotype distributed across 10 vials (20 flies/vial) marked at 90 mm height. Each assay consisted of three trials separated by 15-minute intervals, initiated by sharply tapping vials three times to induce negative geotaxis, with the number of flies reaching the 90 mm mark within 10 seconds recorded per trial. Five groups were tested per genotype (15 trials total). All assays were video recorded under standardized conditions with the camera positioned 20 cm from vials and uniform backlighting provided by a white open-faced box to ensure consistent imaging quality.